# B3_1: odin basics & the SIR model {#sec-lecture1}

**Duration:** ~50 minutes
**Plan:** why odin & the SIR equations (10 min) · first model & compilation
(15) · running & plotting (5) · `user()` & scenarios (10) · cumulative
incidence (5) · exercises (5)

**Goal:** write, compile, run and plot your first compartmental model in odin.

## Why odin?

A compartmental model is a set of ordinary differential equations (ODEs). R's
standard solver, `deSolve`, integrates these with compiled Fortran routines,
so the solver itself is fast. The slow part is the model function: by default
you write it in R, and the solver calls back into R at every integration
step. You can avoid this by writing the model function in C (deSolve supports
this), but doing it by hand is tedious and easy to get wrong.

odin does this translation for you. You write the equations in a language
that looks like R; odin translates them to C, compiles them, and links them
to deSolve's solver. The model then runs at compiled speed, which matters
when you fit models or run many scenarios. odin also handles the bookkeeping
— packing state vectors, initial conditions, parameters and array indexing.

One thing to understand from the start: odin code is declarative. It is not
executed as R code; it is parsed and compiled. You write what the equations
are, not how to compute them, so the order of the lines does not matter.

## The SIR model

The classic SIR model splits a population of $N$ people into:

- $S$ — Susceptible
- $I$ — Infectious
- $R$ — Recovered (removed)

with flows:

$$
\frac{dS}{dt} = -\beta SI, \qquad
\frac{dI}{dt} = \beta SI - \gamma I, \qquad
\frac{dR}{dt} = \gamma I
$$

- $\beta$ — transmission rate
- $\gamma$ — recovery rate; $1/\gamma$ is the mean infectious period
- The basic reproduction number is $R_0 = \beta N / \gamma$

This is the density-dependent formulation: transmission is proportional to
the number of infectious people, so $R_0$ scales with the population size
$N$ — and $\beta$ therefore depends on the population you are modelling.

## Your first odin model

```{r}
library(odin)
```

An odin model has three ingredients: derivatives (`deriv()`), initial
conditions (`initial()`) and parameters. In this first version we fix the
parameter values directly in the code:

```{r}
#| message: true
# verbose = TRUE so we can watch odin generate and compile the C code
sir_generator <- odin::odin({
  ## Derivatives: the flows between compartments
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I

  ## Initial conditions: where the system starts (t = 0)
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0

  ## Parameters: fixed values
  N0    <- 1000  # total population size
  I0    <- 1     # initial number infectious
  beta  <- 2e-4  # transmission rate (R0 = beta * N / gamma = 2)
  gamma <- 0.1   # recovery rate (infectious period of 10 days)
}, verbose = TRUE)
```

The messages show odin generating C code from the equations and compiling it.
From this point on, running the model involves no R code at all. You may also
see compiler warnings (`unused variable`, `format specifies...`). These come
from the C template inside the odin package itself, not from your equations —
odin regenerates that file on every compile, so there is nothing in your
model to correct. They are benign: the unused variable is boilerplate this
small model does not need, and the format-string mismatch sits in an
error-message path that only runs on invalid input. The proper place to fix
them is upstream, in [odin](https://github.com/mrc-ide/odin) itself.

Note two things:

- `odin::odin({...})` parsed and compiled the equations. If that line runs
  without error, the model is ready to use.
- The result, `sir_generator`, is not a model — it is a model generator: a
  factory that creates model instances.

## Running the model

Create an instance and run it over a vector of times:

```{r}
sir <- sir_generator$new()

t <- seq(0, 150, by = 1)
y <- sir$run(t)

head(y)
```

`y` is a data frame with one column per compartment plus time `t`, and one
row per requested time point. Plot it with base R:

```{r}
#| fig-cap: "SIR epidemic curve: R0 = beta * N / gamma = 2"
matplot(y[, "t"], y[, c("S", "I", "R")],
        type = "l", lty = 1, lwd = 2,
        col = c("steelblue", "firebrick", "darkgreen"),
        xlab = "Time (days)", ylab = "Number of people")
legend("right", legend = c("S", "I", "R"), lwd = 2,
       col = c("steelblue", "firebrick", "darkgreen"), bty = "n")
```

## Making parameters changeable: `user()`

This model works, but the parameters are fixed in the compiled C code. To try
a different `beta` you would have to edit the model definition and recompile,
which gets awkward when you want to compare several scenarios.

`user()` solves this. Declaring a parameter with `user(default)` makes it
settable when the model instance is created, with the default used when you
don't set it. Only the parameter block changes:

```{r}
sir_generator <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)  # can now be changed at $new() time
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
})
```

- `sir_generator$new()` uses all the defaults.
- `sir_generator$new(beta = 3e-4)` overrides just `beta`, without
  recompiling — the compiled C code is reused.
- `m$contents()` shows the parameter values an instance is using.

## Changing parameters: scenarios from one compiled model

Scenarios are now cheap — pass new parameter values to `$new()`:

```{r}
#| fig-cap: "Peak prevalence falls as transmission slows"
beta_values <- c(3e-4, 2e-4, 1.5e-4)

plot(NULL, xlim = range(t), ylim = c(0, 350),
     xlab = "Time (days)", ylab = "Infectious (I)")
for (b in beta_values) {
  m <- sir_generator$new(beta = b)
  lines(t, m$run(t)[, "I"], lwd = 2,
        col = hcl.colors(length(beta_values), "Berlin")[which(beta_values == b)])
}
legend("topright", legend = paste("beta =", beta_values,
                                  "(R0 =", beta_values * 1000 / 0.1, ")"),
       col = hcl.colors(length(beta_values), "Berlin"), lwd = 2, bty = "n")
```

Compilation happens once; every scenario after that is just `$new()` and
`$run()`.

## Tracking extra quantities: cumulative incidence

Sometimes you want a quantity that is not a compartment — for example the
total number of infections so far. Add another derivative that accumulates
it:

```{r}
sir_inc <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I   # accumulates new infections

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
})

y <- sir_inc$new()$run(t)
```

```{r}
#| fig-cap: "Cumulative incidence approaches the final epidemic size"
plot(y[, "t"], y[, "cum_inc"], type = "l", lwd = 2,
     xlab = "Time (days)", ylab = "Cumulative infections")
```

The final value of `cum_inc` (≈ `r round(tail(y[, "cum_inc"], 1))` out of
1000) is the final size of the epidemic: about 80% of the population. The
epidemic dies out before everyone is infected because once prevalence is
high, each infectious person infects fewer than one susceptible on average.

::: {.callout-tip collapse="true"}
## Exercise 1.1 — What does $R_0$ control?

Using `sir_generator`, run the model with `beta` chosen so that $R_0$ is 1.2,
1.5 and 2.5 (keep `gamma = 0.1`; remember $R_0 = \beta N / \gamma$). Plot the
three `I` curves on one plot. What happens to the peak height and peak timing
as $R_0$ increases?

```{r}
#| code-fold: true
#| code-summary: "Show solution"
R0_targets <- c(1.2, 1.5, 2.5)
cols <- hcl.colors(length(R0_targets), "Plasma")
plot(NULL, xlim = range(t), ylim = c(0, 250),
     xlab = "Time (days)", ylab = "Infectious (I)")
for (i in seq_along(R0_targets)) {
  b <- R0_targets[i] * 0.1 / 1000     # R0 = beta * N / gamma
  lines(t, sir_generator$new(beta = b)$run(t)[, "I"], col = cols[i], lwd = 2)
}
legend("topright", legend = paste("R0 =", R0_targets),
       col = cols, lwd = 2, bty = "n")
```

Higher $R_0$ gives an earlier, higher, sharper peak; close to the threshold
$R_0 = 1$ the epidemic is slow and shallow.
:::

::: {.callout-tip collapse="true"}
## Exercise 1.2 — Threshold behaviour

Set `beta = 8e-5` (so $R_0 = \beta N / \gamma = 0.8 < 1$) and run the model.
What happens, and why does no epidemic occur?

```{r}
#| code-fold: true
#| code-summary: "Show solution"
y <- sir_generator$new(beta = 8e-5)$run(t)
matplot(y[, "t"], y[, c("S", "I", "R")], type = "l", lty = 1, lwd = 2,
        col = c("steelblue", "firebrick", "darkgreen"),
        xlab = "Time (days)", ylab = "Number of people")
```

When $R_0 < 1$ each infectious person infects fewer than one other on
average, so prevalence decays immediately — `I` never grows.
:::

## Recap

- odin models are equations, not code: `deriv()`, `initial()`, parameters
- `odin::odin({...})` compiles the equations to C and returns a generator
- `generator$new()` creates an instance; `$run(t)` integrates it
- `user(default)` makes a parameter settable at `$new()` — one compilation,
  unlimited scenarios
- Add a `deriv()` line for derived quantities like cumulative incidence

**Next:** @sec-lecture2 — latent periods, births and deaths, and parameters
that change over time.
