# B3_2: Extending models — SEIR, demography & time-varying parameters {#sec-lecture2}

**Goal:** grow the SIR skeleton into models with latent periods, population
dynamics and parameters that change over time.

```{r}
#| include: false
library(odin)
t <- seq(0, 150, by = 1)
```

## Adding a latent period: the SEIR model

For many infections people are exposed but not yet infectious — a latent
period. Add an Exposed compartment $E$; exposed people become infectious at
rate $\sigma$ (so $1/\sigma$ is the mean latent period):

$$
\frac{dS}{dt} = -\beta SI, \quad
\frac{dE}{dt} = \beta SI - \sigma E, \quad
\frac{dI}{dt} = \sigma E - \gamma I, \quad
\frac{dR}{dt} = \gamma I
$$

In odin this means one more compartment and re-routing the flows:

```{r}
seir_generator <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(E) <- beta * S * I - sigma * E
  deriv(I) <- sigma * E - gamma * I
  deriv(R) <- gamma * I

  initial(S) <- N0 - I0
  initial(E) <- 0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)  # R0 = beta * N / gamma = 2
  gamma <- user(0.1)   # infectious period 10 days
  sigma <- user(0.25)  # latent period 4 days
})

y_seir <- seir_generator$new()$run(t)
```

```{r}
#| fig-cap: "SEIR epidemic with a 4-day latent period"
matplot(y_seir[, "t"], y_seir[, c("S", "E", "I", "R")],
        type = "l", lty = 1, lwd = 2,
        col = c("steelblue", "orange", "firebrick", "darkgreen"),
        xlab = "Time (days)", ylab = "Number of people")
legend("right", legend = c("S", "E", "I", "R"), lwd = 2, bty = "n",
       col = c("steelblue", "orange", "firebrick", "darkgreen"))
```

Compared with the SIR from lecture 1, the same $R_0 = \beta N / \gamma$ now
gives a later, broader peak — the latent period slows the epidemic down even
though it does not change $R_0$. Rather than take that on trust, plot both:

```{r}
#| fig-cap: "Same R0 = 2: the latent period delays and lowers the peak"
sir_generator <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
})

y_sir <- sir_generator$new()$run(t)

plot(t, y_sir[, "I"], type = "l", lwd = 2, col = "grey60",
     xlab = "Time (days)", ylab = "Infectious (I)")
lines(t, y_seir[, "I"], lwd = 2, col = "firebrick")
legend("topright", legend = c("SIR", "SEIR (4-day latent period)"),
       col = c("grey60", "firebrick"), lwd = 2, bty = "n")
```

The peak arrives later and is lower — the same total transmission, spread
over a longer time.

## Births and deaths: demography

So far the epidemic is a one-off: everyone ends up in $R$ and nothing else
happens. Over longer horizons, births replenish the susceptible pool. Add a
per-capita birth/death rate $\mu$: births enter $S$, deaths leave every
compartment at rate $\mu$, so $N$ stays constant. We now need
$N = S + E + I + R$ explicitly for the birth term.

To see the long-term behaviour on a short horizon we use fast turnover — a
mean of 2 years in the population (think of a nursery or boarding school
rather than a national population). Slower turnover gives the same
qualitative behaviour, just over decades.

```{r}
seir_demo <- odin::odin({
  deriv(S) <- mu * N - beta * S * I - mu * S
  deriv(E) <- beta * S * I - sigma * E - mu * E
  deriv(I) <- sigma * E - gamma * I - mu * I
  deriv(R) <- gamma * I - mu * R

  N <- S + E + I + R

  initial(S) <- N0 - I0
  initial(E) <- 0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
  sigma <- user(0.25)
  mu    <- user(1 / (2 * 365))  # mean 2 years in the population
})

t_long <- seq(0, 365 * 20, by = 7)   # 20 years, weekly output
y_demo <- seir_demo$new()$run(t_long)
```

```{r}
#| fig-cap: "With births, the infection persists: damped oscillations towards an endemic equilibrium (first 10 years shown)"
plot(y_demo[, "t"] / 365, y_demo[, "I"], type = "l", lwd = 2,
     col = "firebrick", xlab = "Time (years)", ylab = "Infectious (I)",
     xlim = c(0, 10))
```

Instead of dying out after one wave, incidence oscillates with shrinking
amplitude and settles to an endemic equilibrium — newborn susceptibles keep
transmission going. (With the 70-year turnover of a national population, the
first wave would exhaust the susceptibles so thoroughly that the infection
dies out before births can replenish them — worth remembering that these
deterministic models allow unrealistically small prevalence.)

## Time-varying parameters

So far every parameter has been fixed for the whole run. Real drivers change
over time: transmission rises in winter and falls in summer, and it can
shift abruptly when behaviour changes. The solver needs parameter values at
arbitrary times — not just at the times you have data for — so odin takes a
time series of values from you and builds a continuous function from it with
`interpolate()`. We will look at two examples first, then at how the
machinery works.

### A shift in regime: `"constant"`

The simplest case: transmission changes from one level to another at a known
time and stays there — two regimes. With `"constant"` interpolation, `beta`
holds each knot's value until the next knot, which is exactly a step:

```{r}
sir_step <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I

  ## beta as a step function of time
  beta <- interpolate(tt, bb, "constant")

  tt[] <- user()   # regime change times
  bb[] <- user()   # beta in each regime
  dim(tt) <- user()
  dim(bb) <- user()

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})

## two regimes: beta halves at day 40
tt <- c(0, 40, 150)
bb <- c(2e-4, 1e-4, 1e-4)

y_step   <- sir_step$new(tt = tt, bb = bb)$run(t)
y_nostep <- sir_generator$new()$run(t)   # baseline: constant beta
```

(`sir_generator` is the plain SIR from the SEIR comparison.) Plot the
forcing first — always check what you are feeding in:

```{r}
#| fig-cap: "beta as a step function of time"
plot(tt, bb, type = "s", lwd = 2, xlab = "Time (days)",
     ylab = expression(beta), xlim = range(t))
```

```{r}
#| fig-cap: "Halving transmission at day 40 bends the epidemic curve"
plot(t, y_nostep[, "I"], type = "l", lwd = 2, col = "grey60",
     xlab = "Time (days)", ylab = "Infectious (I)")
lines(t, y_step[, "I"], lwd = 2, col = "firebrick")
abline(v = 40, lty = 2)
legend("topright", legend = c("constant beta", "beta halves at day 40"),
       col = c("grey60", "firebrick"), lwd = 2, bty = "n")
```

Note the workflow is unchanged: one compiled model, and the time series is
just another set of parameters supplied at `$new()`.

### Seasonal forcing: `"linear"`

A step is not the right shape for drivers that vary smoothly, like
seasonality. One practical point first: in the closed models so far, a
single wave infects most of the population and then transmission is over —
there is nothing left for seasonality to act on. To see recurrent seasonal
waves we let immunity wane: recovered people return to the susceptible
compartment at rate $\omega$ (mean immunity duration $1/\omega$, here 90
days). This is the SIRS model, with a seasonal `beta` interpolated
linearly:

```{r}
sirs_seasonal <- odin::odin({
  deriv(S) <- -beta * S * I + omega * R
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I - omega * R

  ## beta interpolated from a user-supplied time series
  beta <- interpolate(tt, bb, "linear")

  tt[] <- user()   # time points
  bb[] <- user()   # beta at each time point
  dim(tt) <- user()
  dim(bb) <- user()

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)    # infectious period 10 days
  omega <- user(1 / 90) # immunity lasts 90 days on average
})
```

The forcing series: monthly values over five years, peaking in "winter":

```{r}
t_years <- 5
tt <- seq(0, 365 * t_years, by = 365 / 12)   # monthly knots
bb <- 2e-4 * (1 + 0.3 * cos(2 * pi * tt / 365))
```

```{r}
#| fig-cap: "The seasonal forcing series: monthly knots, linear interpolation in between"
plot(tt / 365, bb, type = "b", pch = 16, cex = 0.5,
     xlab = "Time (years)", ylab = expression(beta))
```

Then pass the vectors at `$new()`; odin infers their lengths:

```{r}
m <- sirs_seasonal$new(tt = tt, bb = bb)
t_run <- seq(0, 365 * t_years, by = 1)
y_seas <- m$run(t_run)
```

```{r}
#| fig-cap: "Seasonal waves: incidence rises and falls with the forced beta"
plot(t_run / 365, y_seas[, "I"], type = "l", lwd = 1.5, col = "firebrick",
     xlab = "Time (years)", ylab = "Infectious (I)")
```

After the first (largest) wave, incidence settles into annual waves that
track the forcing, each peaking shortly after `beta` peaks.

## What `interpolate()` is doing

The solver does not ask for `beta` only at your monthly knots — it evaluates
the equations at whatever times its internal steps require. So odin turns
your points into a continuous function of time: `interpolate(tt, bb, type)`
builds `beta(t)` for *any* `t`, and `type` decides what happens between the
knots:

```{r}
#| fig-cap: "The same monthly knots under the three interpolation types"
#| fig-width: 9
#| fig-height: 5.5
tt_yr <- seq(0, 365, by = 365 / 12)   # one year of knots
bb_yr <- 2e-4 * (1 + 0.3 * cos(2 * pi * tt_yr / 365))
grid  <- seq(0, 365, by = 1)

plot(grid, spline(tt_yr, bb_yr, xout = grid)$y, type = "l", lwd = 2.5,
     col = "#08519c", xlab = "Time (days)", ylab = expression(beta))
lines(grid, approx(tt_yr, bb_yr, xout = grid, method = "linear")$y,
      lwd = 2.5, col = "#d7301f", lty = 2)
lines(grid, approx(tt_yr, bb_yr, xout = grid, method = "constant")$y,
      lwd = 2, col = "grey30")
points(tt_yr, bb_yr, pch = 16, cex = 1.2)
legend("bottomleft",
       legend = c("spline (smooth, solid navy)", "linear (dashed red)",
                  "constant (steps, grey)", "knots"),
       col = c("#08519c", "#d7301f", "grey30", "black"),
       lwd = c(2.5, 2.5, 2, NA), lty = c(1, 2, 1, NA),
       pch = c(NA, NA, NA, 16), bty = "n", cex = 1.1)
```

- `"constant"` holds each value until the next knot — a step function,
  right for quantities that change in steps (e.g. a reporting rate that
  changes month to month)
- `"linear"` joins the dots — a good default for smooth drivers like
  seasonality
- `"spline"` fits a smooth curve through the knots — nicest when the driver
  is genuinely smooth, but it can overshoot between widely spaced knots

A word on the `dim()` declarations. odin compiles your model before it sees
any data, so every array needs a declared size at compile time — there is no
way to leave `dim()` out. Writing `dim(tt) <- user()` means "the length will
be known when the model is instantiated"; at `$new()` you then pass just the
vectors and odin infers their lengths, which is why we never had to state a
number ourselves.

::: {.callout-tip collapse="true"}
## Exercise 2.1 — A less variable latent period

A single $E$ compartment implies an *exponential* latent period: many people
become infectious almost immediately, and a few only after a very long time.
A common trick is to chain two exposed compartments $E_1 \to E_2$, each
progressing at rate `2 * sigma`, which keeps the mean latent period at
`1/sigma` but makes it less variable (an Erlang distribution). Build this
model, run it with `sigma = 1/8` next to the single-$E$ version, and compare
the epidemic curves.

```{r}
#| code-fold: true
#| code-summary: "Show solution"
seir2_generator <- odin::odin({
  deriv(S)  <- -beta * S * I
  deriv(E1) <- beta * S * I - 2 * sigma * E1
  deriv(E2) <- 2 * sigma * E1 - 2 * sigma * E2
  deriv(I)  <- 2 * sigma * E2 - gamma * I
  deriv(R)  <- gamma * I

  initial(S)  <- N0 - I0
  initial(E1) <- 0
  initial(E2) <- 0
  initial(I)  <- I0
  initial(R)  <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
  sigma <- user(1 / 8)  # mean latent period still 8 days
})

y1 <- seir_generator$new(sigma = 1 / 8)$run(t)
y2 <- seir2_generator$new()$run(t)

plot(t, y1[, "I"], type = "l", lwd = 2, col = "grey60",
     xlab = "Time (days)", ylab = "Infectious (I)")
lines(t, y2[, "I"], lwd = 2, col = "firebrick")
legend("topright", legend = c("single E", "chained E1 -> E2"),
       col = c("grey60", "firebrick"), lwd = 2, bty = "n")
```

The curves are very similar — same $R_0$, same final size — but the chained
model takes off later (peak around day 148 vs day 142), because fewer people
race through the latent period unrealistically fast. Chaining more
compartments ($E_1 \to E_2 \to E_3 \dots$) makes the latent period
progressively tighter around its mean.
:::

::: {.callout-tip collapse="true"}
## Exercise 2.2 — Disease-induced mortality

Extend the demography model (`seir_demo`) so that infectious people die at an
*additional* rate `delta` (disease-induced mortality), with the deaths
accumulating in a new compartment `D`. The living population is no longer
constant. Run 20 years with `delta = 0` and `delta = 0.01`, and plot the
total population size over time (compute it in R with `rowSums` on the
compartment columns). What happens?

```{r}
#| code-fold: true
#| code-summary: "Show solution"
seir_mort <- odin::odin({
  deriv(S) <- mu * N - beta * S * I - mu * S
  deriv(E) <- beta * S * I - sigma * E - mu * E
  deriv(I) <- sigma * E - gamma * I - mu * I - delta * I
  deriv(R) <- gamma * I - mu * R
  deriv(D) <- delta * I   # disease-induced deaths accumulate here

  N <- S + E + I + R      # living population (drives births)

  initial(S) <- N0 - I0
  initial(E) <- 0
  initial(I) <- I0
  initial(R) <- 0
  initial(D) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
  sigma <- user(0.25)
  mu    <- user(1 / (2 * 365))  # mean 2 years in the population
  delta <- user(0)   # additional disease-induced mortality rate
})

y0 <- seir_mort$new()$run(t_long)
y1 <- seir_mort$new(delta = 0.01)$run(t_long)

N0_total <- rowSums(y0[, c("S", "E", "I", "R")])
N1_total <- rowSums(y1[, c("S", "E", "I", "R")])
```

First, the plot as many people would draw it by default:

```{r}
#| code-fold: true
#| code-summary: "Show solution"
#| fig-cap: "Default axis limits — a misleading picture"
plot(y0[, "t"] / 365, N0_total, type = "l", lwd = 2, col = "grey60",
     xlab = "Time (years)", ylab = "Total population")
lines(y1[, "t"] / 365, N1_total, lwd = 2, col = "firebrick")
legend("topleft", legend = c("delta = 0", "delta = 0.01"),
       col = c("grey60", "firebrick"), lwd = 2, bty = "n")
```

R picks axis limits from the data range, so the y-axis starts around 725 —
and a loss of about a quarter of the population fills the whole panel and
looks like a collapse. Whenever you plot counts, start the axis at 0:

```{r}
#| code-fold: true
#| code-summary: "Show solution"
#| fig-cap: "Same data, y-axis from 0 — the honest picture"
plot(y0[, "t"] / 365, N0_total, type = "l", lwd = 2, col = "grey60",
     xlab = "Time (years)", ylab = "Total population",
     ylim = c(0, 1000))
lines(y1[, "t"] / 365, N1_total, lwd = 2, col = "firebrick")
legend("bottomleft", legend = c("delta = 0", "delta = 0.01"),
       col = c("grey60", "firebrick"), lwd = 2, bty = "n")
```

Disease-induced deaths accumulate in `D` and the living population declines;
the extra mortality also lowers prevalence slightly because infectious people
are removed faster. And compare the two figures: truncating the y-axis is a
common way to exaggerate an effect — always check where an axis starts.
:::

## Recap

- New compartments = new `deriv()` + `initial()` lines, and re-routing flows
- Births and deaths turn a single epidemic into endemic dynamics
- One generator, many `$new()` calls: the scenario workflow
- Time-varying parameters: `interpolate()` drives a parameter from a
  user-supplied time series — plot the series first, then let odin infer the
  array lengths
