# B3_2: Extending models — SEIR, demography & time-varying parameters {#sec-lecture2}

**Goal:** grow the SIR skeleton into models with latent periods, population
dynamics and parameters that change over time.

```{r}
#| include: false
library(odin)
library(tidyverse)
times <- seq(0, 150, by = 1)   # daily output over 150 days
```

## Adding a latent period: the SEIR model

For many infections people are exposed but not yet infectious — a latent
period. Add an Exposed compartment $E$; exposed people become infectious at
rate $\sigma$ (so $1/\sigma$ is the mean latent period):

$$
\frac{dS}{dt} = -\beta SI, \quad
\frac{dE}{dt} = \beta SI - \sigma E, \quad
\frac{dI}{dt} = \sigma E - \gamma I, \quad
\frac{dR}{dt} = \gamma I
$$

In odin this means one more compartment and re-routing the flows:

```{r}
seir_generator <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(E) <- beta * S * I - sigma * E
  deriv(I) <- sigma * E - gamma * I
  deriv(R) <- gamma * I

  initial(S) <- N0 - I0
  initial(E) <- 0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)  # R0 = beta * N / gamma = 2
  gamma <- user(0.1)   # infectious period 10 days
  sigma <- user(0.25)  # latent period 4 days
})

out_seir <- seir_generator$new()$run(times)   # run with default parameters
```

```{r}
#| fig-cap: "SEIR epidemic with a 4-day latent period"
as.data.frame(out_seir) |>
  pivot_longer(-t, names_to = "compartment", values_to = "n") |>
  mutate(compartment = factor(compartment, levels = c("S", "E", "I", "R"))) |>
  ggplot(aes(t, n, colour = compartment)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    values = c(S = "#0072B2", E = "#E69F00", I = "#D55E00", R = "#009E73")) +
  labs(x = "Time (days)", y = "Number of people", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

Compared with the SIR from lecture 1, the same $R_0 = \beta N / \gamma$ now
gives a later, broader peak — the latent period slows the epidemic down even
though it does not change $R_0$. Rather than take that on trust, plot both:

```{r}
#| fig-cap: "Same R0 = 2: the latent period delays and lowers the peak"
sir_generator <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
})

out_sir <- sir_generator$new()$run(times)   # plain SIR for comparison

tibble(t = times,
       SIR  = out_sir[, "I"],
       `SEIR (4-day latent period)` = out_seir[, "I"]) |>
  pivot_longer(-t, names_to = "model", values_to = "I") |>
  mutate(model = factor(model, levels = c("SIR", "SEIR (4-day latent period)"))) |>
  ggplot(aes(t, I, colour = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    values = c("SIR" = "grey60", "SEIR (4-day latent period)" = "#D55E00")) +
  labs(x = "Time (days)", y = "Infectious (I)", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

The peak arrives later and is lower — the same total transmission, spread
over a longer time.

## Births and deaths: demography

So far the epidemic is a one-off: everyone ends up in $R$ and nothing else
happens. Over longer horizons, births replenish the susceptible pool. Add a
per-capita birth/death rate $\mu$: births enter $S$, deaths leave every
compartment at rate $\mu$, so $N$ stays constant. We now need
$N = S + E + I + R$ explicitly for the birth term.

To see the long-term behaviour on a short horizon we use fast turnover — a
mean of 2 years in the population (think of a nursery or boarding school
rather than a national population). Slower turnover gives the same
qualitative behaviour, just over decades.

```{r}
seir_demo <- odin::odin({
  deriv(S) <- mu * N - beta * S * I - mu * S
  deriv(E) <- beta * S * I - sigma * E - mu * E
  deriv(I) <- sigma * E - gamma * I - mu * I
  deriv(R) <- gamma * I - mu * R

  N <- S + E + I + R

  initial(S) <- N0 - I0
  initial(E) <- 0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
  sigma <- user(0.25)
  mu    <- user(1 / (2 * 365))  # mean 2 years in the population
})

times_long <- seq(0, 365 * 20, by = 7)   # 20 years, weekly output
out_demo <- seir_demo$new()$run(times_long)
```

```{r}
#| fig-cap: "With births, the infection persists: damped oscillations towards an endemic equilibrium (first 10 years shown)"
ggplot(as.data.frame(out_demo), aes(t / 365, I)) +
  geom_line(colour = "#D55E00", linewidth = 1) +
  coord_cartesian(xlim = c(0, 10)) +
  labs(x = "Time (years)", y = "Infectious (I)") +
  theme_classic(base_size = 16)
```

Instead of dying out after one wave, incidence oscillates with shrinking
amplitude and settles to an endemic equilibrium — newborn susceptibles keep
transmission going. (With the 70-year turnover of a national population, the
first wave would exhaust the susceptibles so thoroughly that the infection
dies out before births can replenish them — worth remembering that these
deterministic models allow unrealistically small prevalence.)

## Time-varying parameters

So far every parameter has been fixed for the whole run. Real drivers change
over time: transmission rises in winter and falls in summer, and it can
shift abruptly when behaviour changes. The solver needs parameter values at
arbitrary times — not just at the times you have data for — so odin takes a
time series of values from you and builds a continuous function from it with
`interpolate()`. We will look at two examples first, then at how the
machinery works.

### A shift in regime: `"constant"`

The simplest case: transmission changes from one level to another at a known
time and stays there — two regimes. With `"constant"` interpolation, `beta`
holds each knot's value until the next knot, which is exactly a step:

```{r}
sir_step <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I

  ## beta as a step function of time
  beta <- interpolate(beta_times, beta_vals, "constant")

  beta_times[] <- user()   # regime change times
  beta_vals[] <- user()    # beta in each regime
  dim(beta_times) <- user()
  dim(beta_vals) <- user()

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})

## two regimes: beta halves at day 40
beta_times <- c(0, 40, 150)
beta_vals  <- c(2e-4, 1e-4, 1e-4)

out_step   <- sir_step$new(beta_times = beta_times,
                           beta_vals = beta_vals)$run(times)
out_nostep <- sir_generator$new()$run(times)   # baseline: constant beta
```

(`sir_generator` is the plain SIR from the SEIR comparison.) Plot the
forcing first — always check what you are feeding in:

```{r}
#| fig-cap: "beta as a step function of time"
ggplot(tibble(t = beta_times, beta = beta_vals), aes(t, beta)) +
  geom_step(direction = "hv", linewidth = 1) +
  coord_cartesian(xlim = range(times)) +
  labs(x = "Time (days)", y = expression(beta)) +
  theme_classic(base_size = 16)
```

```{r}
#| fig-cap: "Halving transmission at day 40 bends the epidemic curve"
tibble(t = times,
       `constant beta`         = out_nostep[, "I"],
       `beta halves at day 40` = out_step[, "I"]) |>
  pivot_longer(-t, names_to = "scenario", values_to = "I") |>
  mutate(scenario = factor(scenario,
                           levels = c("constant beta", "beta halves at day 40"))) |>
  ggplot(aes(t, I, colour = scenario)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 40, linetype = "dashed") +
  scale_colour_manual(
    values = c("constant beta" = "grey60", "beta halves at day 40" = "#D55E00")) +
  labs(x = "Time (days)", y = "Infectious (I)", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

Note the workflow is unchanged: one compiled model, and the time series is
just another set of parameters supplied at `$new()`.

### Seasonal forcing: `"linear"`

A step is not the right shape for drivers that vary smoothly, like
seasonality. One practical point first: in the closed models so far, a
single wave infects most of the population and then transmission is over —
there is nothing left for seasonality to act on. To see recurrent seasonal
waves we let immunity wane: recovered people return to the susceptible
compartment at rate $\omega$ (mean immunity duration $1/\omega$, here 90
days). This is the SIRS model, with a seasonal `beta` interpolated
linearly:

```{r}
sirs_seasonal <- odin::odin({
  deriv(S) <- -beta * S * I + omega * R
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I - omega * R

  ## beta interpolated from a user-supplied time series
  beta <- interpolate(beta_times, beta_vals, "linear")

  beta_times[] <- user()   # time points
  beta_vals[] <- user()    # beta at each time point
  dim(beta_times) <- user()
  dim(beta_vals) <- user()

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)    # infectious period 10 days
  omega <- user(1 / 90) # immunity lasts 90 days on average
})
```

The forcing series: monthly values over five years, peaking in "winter":

```{r}
t_years <- 5
beta_times <- seq(0, 365 * t_years, by = 365 / 12)   # monthly knots
beta_vals  <- 2e-4 * (1 + 0.3 * cos(2 * pi * beta_times / 365))
```

```{r}
#| fig-cap: "The seasonal forcing series: monthly knots, linear interpolation in between"
ggplot(tibble(t = beta_times / 365, beta = beta_vals), aes(t, beta)) +
  geom_line() +
  geom_point(shape = 16, size = 1) +
  labs(x = "Time (years)", y = expression(beta)) +
  theme_classic(base_size = 16)
```

Then pass the vectors at `$new()`; odin infers their lengths:

```{r}
mod <- sirs_seasonal$new(beta_times = beta_times, beta_vals = beta_vals)
times_seas <- seq(0, 365 * t_years, by = 1)   # daily output over 5 years
out_seas <- mod$run(times_seas)
```

```{r}
#| fig-cap: "Seasonal waves: incidence rises and falls with the forced beta"
ggplot(as.data.frame(out_seas), aes(t / 365, I)) +
  geom_line(colour = "#D55E00", linewidth = 0.8) +
  labs(x = "Time (years)", y = "Infectious (I)") +
  theme_classic(base_size = 16)
```

After the first (largest) wave, incidence settles into annual waves that
track the forcing, each peaking shortly after `beta` peaks.

## What `interpolate()` is doing

The solver does not ask for `beta` only at your monthly knots — it evaluates
the equations at whatever times its internal steps require. So odin turns
your points into a continuous function of time:
`interpolate(beta_times, beta_vals, type)` builds `beta(t)` for *any* `t`,
and `type` decides what happens between the knots:

```{r}
#| fig-cap: "The same quarterly knots under the three interpolation types"
#| fig-width: 10
#| fig-height: 5.5
knot_times <- seq(0, 365, by = 365 / 4)   # quarterly knots — wider apart on purpose
knot_vals  <- 2e-4 * (1 + 0.3 * cos(2 * pi * knot_times / 365))
days       <- seq(0, 365, by = 1)   # daily points to evaluate the curves on

curves <- tibble(
  day      = days,
  spline   = spline(knot_times, knot_vals, xout = days)$y,
  linear   = approx(knot_times, knot_vals, xout = days, method = "linear")$y,
  constant = approx(knot_times, knot_vals, xout = days, method = "constant")$y
) |>
  pivot_longer(-day, names_to = "method", values_to = "beta") |>
  mutate(method = factor(method, levels = c("spline", "linear", "constant")))

knots <- tibble(day = knot_times, beta = knot_vals)

ggplot(curves, aes(day, beta)) +
  geom_line(aes(colour = method, linetype = method, linewidth = method)) +
  geom_point(data = knots, aes(colour = "knots"), shape = 16, size = 2.5) +
  scale_colour_manual(
    values = c(spline = "#0072B2", linear = "#D55E00",
               constant = "#009E73", knots = "black"),
    breaks = c("spline", "linear", "constant", "knots")) +
  scale_linetype_manual(
    values = c(spline = "solid", linear = "dashed", constant = "solid"),
    guide = "none") +
  scale_linewidth_manual(
    values = c(spline = 1.2, linear = 1.2, constant = 1), guide = "none") +
  guides(colour = guide_legend(override.aes = list(
    linetype = c("solid", "dashed", "solid", "blank"),
    shape    = c(NA, NA, NA, 16)))) +
  labs(x = "Time (days)", y = expression(beta), colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

With quarterly knots the difference is visible: `"linear"` is a triangle
wave through the points, `"spline"` bends smoothly — and overshoots the
knot values on both sides. With monthly knots (as in the seasonal example)
linear and spline are nearly indistinguishable.

- `"constant"` holds each value until the next knot — a step function,
  right for quantities that change in steps (e.g. a reporting rate that
  changes month to month)
- `"linear"` joins the dots — a good default for smooth drivers like
  seasonality
- `"spline"` fits a smooth curve through the knots — nicest when the driver
  is genuinely smooth, but it can overshoot between widely spaced knots

A word on the `dim()` declarations. odin compiles your model before it sees
any data, so every array needs a declared size at compile time — there is no
way to leave `dim()` out. Writing `dim(beta_times) <- user()` means "the
length will be known when the model is instantiated"; at `$new()` you then
pass just the vectors and odin infers their lengths, which is why we never
had to state a number ourselves.

::: {.callout-tip collapse="true"}
## Exercise 2.1 — A less variable latent period

A single $E$ compartment implies an *exponential* latent period: the most
likely outcome is becoming infectious **immediately**, with a long tail of
very long stays. Its variance equals the mean squared — the most variable
distribution possible for a given mean, and not what real latent periods
look like.

A common trick is to chain two exposed compartments $E_1 \to E_2$, each
progressing at rate `2 * sigma`. The mean latent period is still `1/sigma`
(two stages of mean `1/(2*sigma)` each), but the distribution becomes
Erlang (Gamma with shape 2): the variance is **halved**, so latent periods
cluster around the mean instead of piling up near zero. Build this model,
run it with `sigma = 1/8` next to the single-$E$ version, and compare the
epidemic curves.

```{r}
#| code-fold: true
#| code-summary: "Show solution"
seir2_generator <- odin::odin({
  deriv(S)  <- -beta * S * I
  deriv(E1) <- beta * S * I - 2 * sigma * E1
  deriv(E2) <- 2 * sigma * E1 - 2 * sigma * E2
  deriv(I)  <- 2 * sigma * E2 - gamma * I
  deriv(R)  <- gamma * I

  initial(S)  <- N0 - I0
  initial(E1) <- 0
  initial(E2) <- 0
  initial(I)  <- I0
  initial(R)  <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
  sigma <- user(1 / 8)  # mean latent period still 8 days
})

out_single  <- seir_generator$new(sigma = 1 / 8)$run(times)
out_chained <- seir2_generator$new()$run(times)

tibble(t = times,
       `single E`         = out_single[, "I"],
       `chained E1 -> E2` = out_chained[, "I"]) |>
  pivot_longer(-t, names_to = "model", values_to = "I") |>
  mutate(model = factor(model, levels = c("single E", "chained E1 -> E2"))) |>
  ggplot(aes(t, I, colour = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    values = c("single E" = "grey60", "chained E1 -> E2" = "#D55E00")) +
  labs(x = "Time (days)", y = "Infectious (I)", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

The curves are very similar — same $R_0$, same final size — but the chained
model takes off later (peak around day 148 vs day 142), because fewer people
race through the latent period unrealistically fast. Chaining more
compartments ($E_1 \to E_2 \to E_3 \dots$) makes the latent period
progressively tighter around its mean.

To see *why* the difference is small here, look at the two latent-period
distributions the models imply:

```{r}
#| code-fold: true
#| code-summary: "Show solution"
#| fig-cap: "Same 8-day mean latent period: exponential (single E) vs Erlang-2 (chained)"
latent_days <- seq(0, 30, by = 0.1)   # latent periods to evaluate the density on
tibble(day = latent_days,
       `single E: exponential, variance 64` = dexp(latent_days, rate = 1 / 8),
       `chained: Erlang-2, variance 32` =
         dgamma(latent_days, shape = 2, rate = 2 / 8)) |>
  pivot_longer(-day, names_to = "model", values_to = "density") |>
  mutate(model = factor(model,
                        levels = c("single E: exponential, variance 64",
                                   "chained: Erlang-2, variance 32"))) |>
  ggplot(aes(day, density, colour = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    values = c("single E: exponential, variance 64" = "grey60",
               "chained: Erlang-2, variance 32" = "#D55E00")) +
  labs(x = "Latent period (days)", y = "Density", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

The grey exponential says the most likely latent period is near zero — many
people become infectious almost immediately. The red Erlang-2 is zero at
zero, peaks near the mean, and has half the variance: latent periods cluster
around 8 days. Same mean — what changes is the *spread*, and that is what
the extra compartment buys you.
:::

::: {.callout-tip collapse="true"}
## Exercise 2.2 — Disease-induced mortality

Extend the demography model (`seir_demo`) so that infectious people die at an
*additional* rate `delta` (disease-induced mortality), with the deaths
accumulating in a new compartment `D`. The living population is no longer
constant. Run 20 years with `delta = 0` and `delta = 0.01`, and plot the
total population size over time (compute it in R with `rowSums` on the
compartment columns). What happens?

```{r}
#| code-fold: true
#| code-summary: "Show solution"
seir_mort <- odin::odin({
  deriv(S) <- mu * N - beta * S * I - mu * S
  deriv(E) <- beta * S * I - sigma * E - mu * E
  deriv(I) <- sigma * E - gamma * I - mu * I - delta * I
  deriv(R) <- gamma * I - mu * R
  deriv(D) <- delta * I   # disease-induced deaths accumulate here

  N <- S + E + I + R      # living population (drives births)

  initial(S) <- N0 - I0
  initial(E) <- 0
  initial(I) <- I0
  initial(R) <- 0
  initial(D) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
  sigma <- user(0.25)
  mu    <- user(1 / (2 * 365))  # mean 2 years in the population
  delta <- user(0)   # additional disease-induced mortality rate
})

out_nomort <- seir_mort$new()$run(times_long)
out_mort   <- seir_mort$new(delta = 0.01)$run(times_long)

pop_nomort <- rowSums(out_nomort[, c("S", "E", "I", "R")])
pop_mort   <- rowSums(out_mort[, c("S", "E", "I", "R")])
```

First, the plot as many people would draw it by default:

```{r}
#| code-fold: true
#| code-summary: "Show solution"
#| fig-cap: "Default axis limits — a misleading picture"
tibble(t = out_nomort[, "t"] / 365,
       `delta = 0`    = pop_nomort,
       `delta = 0.01` = pop_mort) |>
  pivot_longer(-t, names_to = "scenario", values_to = "pop") |>
  mutate(scenario = factor(scenario, levels = c("delta = 0", "delta = 0.01"))) |>
  ggplot(aes(t, pop, colour = scenario)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    values = c("delta = 0" = "grey60", "delta = 0.01" = "#D55E00")) +
  labs(x = "Time (years)", y = "Total population", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

R picks axis limits from the data range, so the y-axis starts around 725 —
and a loss of about a quarter of the population fills the whole panel and
looks like a collapse. Whenever you plot counts, start the axis at 0:

```{r}
#| code-fold: true
#| code-summary: "Show solution"
#| fig-cap: "Same data, y-axis from 0 — the honest picture"
tibble(t = out_nomort[, "t"] / 365,
       `delta = 0`    = pop_nomort,
       `delta = 0.01` = pop_mort) |>
  pivot_longer(-t, names_to = "scenario", values_to = "pop") |>
  mutate(scenario = factor(scenario, levels = c("delta = 0", "delta = 0.01"))) |>
  ggplot(aes(t, pop, colour = scenario)) +
  geom_line(linewidth = 1) +
  coord_cartesian(ylim = c(0, 1000)) +
  scale_colour_manual(
    values = c("delta = 0" = "grey60", "delta = 0.01" = "#D55E00")) +
  labs(x = "Time (years)", y = "Total population", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

Disease-induced deaths accumulate in `D` and the living population declines;
the extra mortality also lowers prevalence slightly because infectious people
are removed faster. And compare the two figures: truncating the y-axis is a
common way to exaggerate an effect — always check where an axis starts.
:::

## Recap

- New compartments = new `deriv()` + `initial()` lines, and re-routing flows
- Births and deaths turn a single epidemic into endemic dynamics
- One generator, many `$new()` calls: the scenario workflow
- Time-varying parameters: `interpolate()` drives a parameter from a
  user-supplied time series — plot the series first, then let odin infer the
  array lengths
