# Estimating time-varying beta {.unnumbered}

This chapter is a set of personal working notes. It is built only in the
local version of the book and is not published.

```{r}
#| include: false
library(odin)
library(tidyverse)
library(patchwork)
set.seed(42)
times <- seq(0, 150, by = 1)   # daily output over 150 days
```

## Three things named "beta"

The word "beta" gets used for three different things, and keeping them
apart is half the battle:

- **A model parameter** — one number per run:
  `mod <- sir_inc$new(beta = 2e-4)`.
- **A scenario vector** — many runs, one per value:
  `beta_values <- c(3e-4, 2e-4, 1.5e-4)`, then `lapply` over it.
- **A paired time series** — `beta_times` and `beta_vals` fed to
  `interpolate()`, so that beta varies *within* one run.

This chapter is about the third: how to specify a time-varying beta, and
how to estimate one from data. All the examples use the same closed SIR
model, with one addition — a `cum_inc` compartment that accumulates new
infections, because fitting needs incidence and surveillance reports
incidence, not prevalence:

```{r}
sir_inc <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I   # accumulates new infections

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
})
```

For the time series we need a variant where beta comes from
`interpolate()`. With `"constant"` interpolation beta holds each knot's
value until the next knot, which is exactly a step:

```{r}
sir_step <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I

  beta <- interpolate(beta_times, beta_vals, "constant")

  beta_times[] <- user()   # regime change times
  beta_vals[] <- user()    # beta in each regime
  dim(beta_times) <- user()
  dim(beta_vals) <- user()

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})
```

The three uses side by side:

```{r}
## (1) one constant beta per run
out <- sir_inc$new(beta = 2e-4)$run(times)

## (2) a scenario vector: one run per value
beta_values <- c(3e-4, 2e-4, 1.5e-4)
scenario_runs <- lapply(beta_values,
                        function(beta_i) sir_inc$new(beta = beta_i)$run(times))

## (3) a paired time series: beta varies within one run
out_step <- sir_step$new(beta_times = c(0, 40, 150),
                         beta_vals = c(2e-4, 1e-4, 1e-4))$run(times)
```

Use (3) is the one that looks least like the others, so it is worth
seeing directly: the top panel is the beta(t) the solver actually sees,
the bottom panel is the epidemic curve from the same single run.

```{r}
#| fig-cap: "One run with a time-varying beta: the step in beta (top) bends the incidence curve (bottom)"
## top: the beta(t) the solver sees, built from the pair
beta_df <- data.frame(time = c(0, 40, 150), beta = c(2e-4, 1e-4, 1e-4))
inc_df <- as.data.frame(out_step)

p1 <- ggplot(beta_df, aes(x = time, y = beta)) +
  geom_step(linewidth = 1.2) +
  geom_vline(xintercept = 40, linetype = "dashed", colour = "grey60") +
  labs(x = NULL, y = expression(beta)) +
  theme_classic(base_size = 16) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

## bottom: incidence from the same single run
p2 <- ggplot(inc_df, aes(x = t, y = I)) +
  geom_line(linewidth = 1.2, colour = "#D55E00") +
  geom_vline(xintercept = 40, linetype = "dashed", colour = "grey60") +
  labs(x = "Time (days)", y = "Infectious (I)") +
  theme_classic(base_size = 16)

p1 / p2
```

`beta_times` says *when*, `beta_vals` says *what*. Everything else in
this chapter is about working backwards from data to that pair.

## Estimating a constant beta

Start with the simplest estimation problem: beta does not vary at all,
and we want to recover it from noisy weekly reports. Simulate the
"observed" data by running the true model and adding Poisson noise to
the weekly incidence:

```{r}
week_times <- seq(0, 150, by = 7)            # weekly output
out <- sir_inc$new()$run(week_times)         # truth: beta = 2e-4
inc_true <- diff(out[, "cum_inc"])           # true weekly incidence
obs <- rpois(length(inc_true), inc_true)     # noisy "reported" data
```

```{r}
#| fig-cap: "The fake surveillance data: noisy weekly incidence (points) around the truth (line)"
surv_df <- data.frame(time = week_times[-1], observed = obs, true = inc_true)
ggplot(surv_df, aes(x = time)) +
  geom_point(aes(y = observed), shape = 16) +
  geom_line(aes(y = true), colour = "#D55E00", linewidth = 1) +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

The error function runs the model at a candidate beta and sums the
squared differences between modelled and observed weekly incidence:

```{r}
sse <- function(beta) {
  out <- sir_inc$new(beta = beta)$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs)^2)
}

## deliberately broad: R0 anywhere from 0.01 to 10
fit <- optimize(sse, interval = c(1e-6, 1e-3), tol = 1e-8)
fit$minimum   # fitted beta; truth is 2e-4
```

The fitted value is about 2.00e-4 — spot on, from an interval that knows
nothing about the truth. The `tol = 1e-8` is not decorative:
`optimize()`'s default tolerance is about 1e-4, which is as big as beta
itself. With the default, the algorithm stops once the bracket is
smaller than that, "converging" to 2.2e-4 — a visibly wrong curve that
peaks too early and too high. Whenever the parameter you are minimising
over is much smaller than 1, set `tol` explicitly.

```{r}
#| fig-cap: "The fitted curve (red) is nearly indistinguishable from the truth (grey)"
out_fit <- sir_inc$new(beta = fit$minimum)$run(week_times)
fit_df <- data.frame(time = week_times[-1], observed = obs, true = inc_true,
                     fitted = diff(out_fit[, "cum_inc"]))
fit_long <- tidyr::pivot_longer(fit_df, -time,
                                names_to = "series", values_to = "incidence")
ggplot(fit_long, aes(x = time, y = incidence, colour = series)) +
  geom_line(data = filter(fit_long, series != "observed"), linewidth = 1) +
  geom_point(data = filter(fit_long, series == "observed"), shape = 16) +
  scale_colour_manual(values = c(observed = "black", true = "grey60",
                                 fitted = "#D55E00")) +
  labs(x = "Time (days)", y = "Weekly new infections", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

A few practical notes before moving to harder problems:

- `optimize()` is 1D only. For several parameters use
  `optim(pars, fn, method = "L-BFGS-B", lower = ..., upper = ...)`.
- Bounds can be genuinely broad — here beta ranged over three orders of
  magnitude. You do not need to know the neighbourhood of the truth; you
  need broad bounds and, once there is more than one parameter, several
  starts.
- Parameters on very different scales (beta ~ 1e-4, a time ~ 40) confuse
  the default finite-difference steps in `optim()` — set
  `control = list(parscale = ...)` with a typical size for each
  parameter.
- Least squares on counts is a stand-in for a Poisson likelihood. It is
  fine for exploring and for point estimates, but it gives no intervals.

## Estimating a regime change

Now let beta change once: transmission halves at day 40. The observed
data come from the step model with `beta_times = c(0, 40, 150)` and
`beta_vals = c(2e-4, 1e-4, 1e-4)`:

```{r}
out_step <- sir_step$new(beta_times = c(0, 40, 150),
                         beta_vals = c(2e-4, 1e-4, 1e-4))$run(week_times)
inc_step <- diff(out_step[, "cum_inc"])
obs_step <- rpois(length(inc_step), inc_step)
```

```{r}
#| fig-cap: "Weekly incidence when transmission halves at day 40 (dashed line)"
step_df <- data.frame(time = week_times[-1], observed = obs_step)
ggplot(step_df, aes(x = time, y = observed)) +
  geom_point(shape = 16) +
  geom_vline(xintercept = 40, linetype = "dashed") +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

We fit three parameters: the two levels and the change time. The error
function rebuilds the interpolate inputs from the parameters — note how
`beta_times` is constructed around the candidate `t_change`:

```{r}
sse_step <- function(pars) {   # pars = c(beta1, beta2, t_change)
  out <- sir_step$new(beta_times = c(0, pars[3], 150),
                      beta_vals = c(pars[1], pars[2], pars[2]))$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs_step)^2)
}
```

With three parameters the error surface has local minima, and with broad
bounds a single start can land in one. The honest workflow is a small
grid of starts spread across the range, keeping whichever end point has
the lowest final error:

```{r}
starts <- rbind(c(1e-4, 1e-4, 25), c(8e-4, 8e-4, 120),
                c(8e-4, 1e-5, 75), c(1e-5, 8e-4, 100))
fits <- lapply(seq_len(nrow(starts)), function(i)
  optim(starts[i, ], sse_step, method = "L-BFGS-B",
        lower = c(1e-6, 1e-6, 5), upper = c(1e-3, 1e-3, 145),
        control = list(parscale = c(1e-4, 1e-4, 10))))
sapply(fits, "[[", "value")   # final SSE per start
```

Two of the four starts converge to the right place (final error about
142); the other two land in wrong minima with about 40 times worse
error. This is the local-optima lesson in its most useful form: the
failures are not subtle — comparing final SSE across starts, rather than
trusting the first answer, is what saves you.

```{r}
fit_step <- fits[[which.min(sapply(fits, "[[", "value"))]]
fit_step$par   # truth: 2e-4, 1e-4, day 40
```

The best fit recovers beta1 = 2.003e-4, beta2 = 9.83e-5 and a change at
day 42.4, against true values of 2e-4, 1e-4 and day 40. The change time
is a little late because the weekly data only pin the bend down to
within a reporting interval or so.

```{r}
#| fig-cap: "Fitted step model (line) against the noisy data; dashed line at the fitted change time"
out_fitstep <- sir_step$new(
  beta_times = c(0, fit_step$par[3], 150),
  beta_vals = c(fit_step$par[1], fit_step$par[2], fit_step$par[2]))$run(week_times)
fitstep_df <- data.frame(time = week_times[-1], observed = obs_step,
                         fitted = diff(out_fitstep[, "cum_inc"]))
ggplot(fitstep_df, aes(x = time)) +
  geom_point(aes(y = observed), shape = 16) +
  geom_line(aes(y = fitted), colour = "#D55E00", linewidth = 1) +
  geom_vline(xintercept = fit_step$par[3], linetype = "dashed") +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

One caveat on identifiability: the change time is only learnable if the
curve visibly bends. If the two regimes produce similar incidence over
the observed window, the data cannot say when the change happened — no
optimizer can fix that.

## Estimating seasonal beta

For a driver that varies smoothly, like seasonality, a step is the wrong
shape. And a practical point first: in the closed SIR a single wave
infects most of the population and then transmission is over — there is
nothing left for seasonality to act on. So for this example we let
immunity wane (the SIRS model, mean immunity 90 days) and drive beta
with a cosine, interpolated linearly between monthly knots:

```{r}
sirs_seasonal <- odin::odin({
  deriv(S) <- -beta * S * I + omega * R
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I - omega * R
  deriv(cum_inc) <- beta * S * I

  beta <- interpolate(beta_times, beta_vals, "linear")

  beta_times[] <- user()
  beta_vals[] <- user()
  dim(beta_times) <- user()
  dim(beta_vals) <- user()

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
  omega <- user(1 / 90)
})
```

The observed data: two years of weekly reports, with transmission
peaking in mid-January (phase = day 15) at 30% amplitude:

```{r}
beta_times <- seq(0, 730, by = 365 / 12)   # monthly knots, 2 years
beta_vals  <- 2e-4 * (1 + 0.3 * cos(2 * pi * (beta_times - 15) / 365))
times_seas <- seq(0, 730, by = 7)          # weekly output
out_seas <- sirs_seasonal$new(beta_times = beta_times,
                              beta_vals = beta_vals)$run(times_seas)
obs_seas <- rpois(length(times_seas) - 1, diff(out_seas[, "cum_inc"]))
```

```{r}
#| fig-cap: "Two years of noisy weekly incidence under seasonal forcing"
seas_df <- data.frame(time = times_seas[-1], observed = obs_seas)
ggplot(seas_df, aes(x = time, y = observed)) +
  geom_point(shape = 16, size = 1) +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

The error function constructs the seasonal beta(t) from three
parameters — baseline, amplitude and phase — on the same monthly knots:

```{r}
sse_seas <- function(pars) {   # pars = c(beta0, amplitude, phase)
  bb <- pars[1] * (1 + pars[2] * cos(2 * pi * (beta_times - pars[3]) / 365))
  out <- sirs_seasonal$new(beta_times = beta_times,
                           beta_vals = bb)$run(times_seas)
  sum((diff(out[, "cum_inc"]) - obs_seas)^2)
}
```

Phase is the hardest of the three to fit: the timing of the peak is only
visible where incidence turns, and the error surface has a ridge with
poor minima on both sides. Three starts, all far from the truth and
spread over phase:

```{r}
starts <- rbind(c(5e-4, 0.5, 100), c(5e-4, 0.5, 180), c(5e-4, 0.5, 300))
fits <- lapply(seq_len(nrow(starts)), function(i)
  optim(starts[i, ], sse_seas, method = "L-BFGS-B",
        lower = c(1e-6, 0, 0), upper = c(1e-3, 1, 365),
        control = list(parscale = c(1e-4, 0.1, 30))))
sapply(fits, "[[", "value")
```

The three outcomes are instructive. The start at phase 100 drifts to the
year boundary (phase 365, error about 8200); the start at phase 180
falls into a flat-beta minimum with amplitude 0 (error about 64000 —
the optimizer decides there is no seasonality at all); only the start at
phase 300 finds the truth. Best of the grid:

```{r}
fit_seas <- fits[[which.min(sapply(fits, "[[", "value"))]]
fit_seas$par   # truth: 2e-4, 0.3, 15
```

The winning fit recovers beta0 = 2.0026e-4, amplitude 0.300 and phase
15.4 days, against true values of 2e-4, 0.3 and 15. When the answer
matters, more seasons of data help more than denser sampling within one
season — each extra winter is another vote for where the peak sits.

```{r}
#| fig-cap: "The fitted seasonal model (line) against the noisy data"
bb_fit <- fit_seas$par[1] *
  (1 + fit_seas$par[2] * cos(2 * pi * (beta_times - fit_seas$par[3]) / 365))
out_fitseas <- sirs_seasonal$new(beta_times = beta_times,
                                 beta_vals = bb_fit)$run(times_seas)
fitseas_df <- data.frame(time = times_seas[-1], observed = obs_seas,
                         fitted = diff(out_fitseas[, "cum_inc"]))
ggplot(fitseas_df, aes(x = time)) +
  geom_point(aes(y = observed), shape = 16, size = 1) +
  geom_line(aes(y = fitted), colour = "#D55E00", linewidth = 0.75) +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

## Covariate-driven beta

Sometimes the shape of beta(t) is not estimated at all — it is measured.
Mobility indices, school term calendars and temperature series all enter
transmission models this way: as a known covariate that scales beta.
Make a concrete example — a weekly "mobility index" that dips around
day 60 and recovers, normalised to mean 1, so that
beta(t) = beta0 x driver(t):

```{r}
#| fig-cap: "Left: the driver series (mean 1). Right: the resulting beta(t) with beta0 = 2e-4."
driver_times <- seq(0, 150, by = 7)
dip <- 1 - 0.6 * exp(-((driver_times - 60)^2) / (2 * 20^2))  # dip at day 60
driver <- dip / mean(dip)            # rescale to mean 1
beta_vals <- 2e-4 * driver           # beta(t) = beta0 * driver(t)
driver_df <- data.frame(time = driver_times, driver = driver, beta = beta_vals)

p1 <- ggplot(driver_df, aes(x = time, y = driver)) +
  geom_line(linewidth = 1) +
  labs(x = "Time (days)", y = "Mobility index") +
  theme_classic(base_size = 16)

p2 <- ggplot(driver_df, aes(x = time, y = beta)) +
  geom_line(linewidth = 1, colour = "#D55E00") +
  labs(x = "Time (days)", y = expression(beta(t))) +
  theme_classic(base_size = 16)

p1 | p2
```

Passing `driver_times` and `beta_vals` to the model is all the
specification takes. The interesting inference question is the
rescaling: if the driver's *shape* is trusted but its *scale* is not,
fit beta0. For that we use a linear-interpolation variant of the SIR, so
beta follows the driver smoothly between the weekly points:

```{r}
sir_driver <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I

  beta <- interpolate(beta_times, beta_vals, "linear")

  beta_times[] <- user()   # driver time points
  beta_vals[] <- user()    # beta at each time point
  dim(beta_times) <- user()
  dim(beta_vals) <- user()

  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0

  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})
```

Simulate observed data with a true scaling of 3e-4, then fit only
beta0, treating the driver as a known covariate inside the error
function:

```{r}
out_d <- sir_driver$new(beta_times = driver_times,
                        beta_vals = 3e-4 * driver)$run(week_times)   # truth
obs_driver <- rpois(length(week_times) - 1, diff(out_d[, "cum_inc"]))

sse_driver <- function(beta0) {   # driver known; fit only its scale
  out <- sir_driver$new(beta_times = driver_times,
                        beta_vals = beta0 * driver)$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs_driver)^2)
}
fit_driver <- optimize(sse_driver, interval = c(1e-6, 1e-3), tol = 1e-8)
fit_driver$minimum   # truth: 3e-4
```

The fitted scaling is 2.99e-4 against a truth of 3e-4, and the fitted
beta(t) lies on top of the true one:

```{r}
#| fig-cap: "True beta(t) (solid) and the fitted scaling (dashed) — nearly identical"
beta_df <- data.frame(time = driver_times,
                      `true beta(t)` = 3e-4 * driver,
                      `fitted beta(t)` = fit_driver$minimum * driver,
                      check.names = FALSE)
beta_long <- tidyr::pivot_longer(beta_df, -time,
                                 names_to = "series", values_to = "beta")
ggplot(beta_long, aes(x = time, y = beta, colour = series, linetype = series)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("true beta(t)" = "black",
                                 "fitted beta(t)" = "#D55E00")) +
  scale_linetype_manual(values = c("true beta(t)" = "solid",
                                   "fitted beta(t)" = "dashed")) +
  coord_cartesian(ylim = c(0, 4e-4)) +
  labs(x = "Time (days)", y = expression(beta(t)),
       colour = NULL, linetype = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

This pattern generalises directly:

- Covariates enter transmission models exactly like this: known series
  in, one coefficient out.
- With several covariates, fit one coefficient each, all together.
- Normalise or centre each covariate first so the coefficients are
  interpretable — here beta0 is the average beta because the driver has
  mean 1.

## Free-form beta(t)

The most flexible option is to give up on a parametric shape and
estimate beta at a set of knots directly — here six, on the constant-beta
data from the first example, where the truth is a flat line:

```{r}
knot_times <- seq(0, 150, by = 30)   # six free betas on the Part-2 data
sse_free <- function(betas) {
  out <- sir_step$new(beta_times = knot_times,
                      beta_vals = betas)$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs)^2)
}
fit_free <- optim(rep(2e-4, 6), sse_free, method = "L-BFGS-B",
                  lower = rep(0, 6), upper = rep(5e-4, 6),
                  control = list(parscale = rep(1e-4, 6), maxit = 2000))
fit_free$par   # truth is a flat 2e-4
```

```{r}
#| fig-cap: "Six freely fitted betas (step line) around the true constant value (dashed) — the wiggle is overfitting"
free_df <- data.frame(time = knot_times, beta = fit_free$par)
ggplot(free_df, aes(x = time, y = beta)) +
  geom_step(linewidth = 1) +
  geom_point(shape = 16) +
  geom_hline(yintercept = 2e-4, colour = "#D55E00", linetype = "dashed") +
  coord_cartesian(xlim = c(0, 150), ylim = c(0, 4e-4)) +
  labs(x = "Time (days)", y = expression(beta)) +
  theme_classic(base_size = 16)
```

The fitted values range from 1.96e-4 to 2.19e-4: the truth is flat, but
the fit wiggles by about 10% in both directions, chasing noise. With one
parameter per knot, the model has enough freedom to fit the noise as
well as the signal. The standard fix is to penalise roughness — add a
term like `lambda * sum(diff(betas)^2)` to the error so that
neighbouring betas are pulled towards each other. Once you want proper
uncertainty intervals on a flexible beta(t), least squares has run out
of road: that is the point to move to odin2 + monty, where priors do the
smoothing and you get full Bayesian inference.

## Choosing an approach

- **Regime change** (policy, behaviour shift): `"constant"`
  interpolation, fit the levels and the change times — broad bounds,
  several starts.
- **Seasonal**: parametric cosine — fit beta0, amplitude and phase;
  spread the starts over phase.
- **Measured driver**: pass the series as `beta_times` / `beta_vals`
  and fit a single scaling coefficient.
- **Unknown shape**: free knots plus a smoothing penalty; if you need
  intervals, move to odin2 + monty.

In every case: plot the series you feed in, and plot the fit against the
data before believing any number.

---

[Slides: estimating and specifying time-varying beta](time-varying-beta.html)
