---
pagetitle: "Estimating and specifying time-varying beta"
format:
  revealjs:
    theme:
      - default
      - ../theme/slides.scss
    smaller: true
    slide-number: true
    embed-resources: true
echo: true
---

```{r}
#| include: false
library(odin)
library(tidyverse)
library(patchwork)
set.seed(42)
times <- seq(0, 150, by = 1)   # daily output over 150 days

## interpolate() variant of the SIR, used in Part 1 and defined in full
## in Part 3
sir_step <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I
  beta <- interpolate(beta_times, beta_vals, "constant")
  beta_times[] <- user()
  beta_vals[] <- user()
  dim(beta_times) <- user()
  dim(beta_vals) <- user()
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})
```

## Estimating and specifying time-varying beta {.title-slide}

Personal notes — not for distribution

## Three different things named "beta"

- **A model parameter** — one number per run:
  `mod <- sir_inc$new(beta = 2e-4)`
- **A scenario vector** — many runs, one per value:
  `beta_values <- c(3e-4, 2e-4, 1.5e-4)`, then `lapply`
- **A paired time series** — `beta_times` / `beta_vals` fed to
  `interpolate()`: beta varies *within* one run

The third is what these notes are about: how to specify it, and how to
estimate it from data.

## The workhorse: SIR with cumulative incidence

For fitting we need incidence, so add a `cum_inc` compartment:

```{r}
sir_inc <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I   # accumulates new infections
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
})
```

## The three uses in code

```{r}
## (1) one constant beta per run
out <- sir_inc$new(beta = 2e-4)$run(times)

## (2) a scenario vector: one run per value
beta_values <- c(3e-4, 2e-4, 1.5e-4)
scenario_runs <- lapply(beta_values,
                        function(beta_i) sir_inc$new(beta = beta_i)$run(times))

## (3) a paired time series: beta varies within one run
out_step <- sir_step$new(beta_times = c(0, 40, 150),
                         beta_vals = c(2e-4, 1e-4, 1e-4))$run(times)
```

## One run, beta varying in time

```{r}
#| fig-height: 5.5
## top: the beta(t) the solver sees, built from the pair
beta_df <- tibble(time = c(0, 40, 150), beta = c(2e-4, 1e-4, 1e-4))
p_beta <- ggplot(beta_df, aes(time, beta)) +
  geom_step(linewidth = 1.25) +
  geom_vline(xintercept = 40, linetype = "dashed", colour = "grey60") +
  labs(x = NULL, y = expression(beta)) +
  theme_classic(base_size = 16)
## bottom: incidence from the same single run
inc_df <- as.data.frame(out_step)
p_inc <- ggplot(inc_df, aes(t, I)) +
  geom_line(colour = "#D55E00", linewidth = 1.25) +
  geom_vline(xintercept = 40, linetype = "dashed", colour = "grey60") +
  labs(x = "Time (days)", y = "Infectious (I)") +
  theme_classic(base_size = 16)
p_beta / p_inc
```

One simulation: `beta` halves at day 40 (top) and the epidemic curve bends
at exactly that point (bottom) — `beta_times` says *when*, `beta_vals` says
*what*.

## Part 2: fake data with a known beta

```{r}
week_times <- seq(0, 150, by = 7)            # weekly output
out <- sir_inc$new()$run(week_times)         # truth: beta = 2e-4
inc_true <- diff(out[, "cum_inc"])           # true weekly incidence
obs <- rpois(length(inc_true), inc_true)     # noisy "reported" data
```

```{r}
#| fig-height: 3.6
obs_df <- tibble(time = week_times[-1], obs = obs, inc_true = inc_true)
ggplot(obs_df, aes(time)) +
  geom_point(aes(y = obs), shape = 16) +
  geom_line(aes(y = inc_true), colour = "#D55E00", linewidth = 1) +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

## Least squares against the model

```{r}
sse <- function(beta) {
  out <- sir_inc$new(beta = beta)$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs)^2)
}

## deliberately broad: R0 anywhere from 0.01 to 10
fit <- optimize(sse, interval = c(1e-6, 1e-3), tol = 1e-8)
fit$minimum   # fitted beta; truth is 2e-4
```

Spot on, from an interval that knows nothing about the truth. But watch
the tolerance: `optimize()`'s default `tol` is about 1e-4 — as big as
beta itself — so without `tol = 1e-8` it stops at 2.2e-4, a visibly
wrong curve.

## The fitted curve

```{r}
#| fig-height: 4.6
out_fit <- sir_inc$new(beta = fit$minimum)$run(week_times)
fit_df <- tibble(
  time = week_times[-1],
  observed = obs,
  true = inc_true,
  fitted = diff(out_fit[, "cum_inc"])
)
ggplot(fit_df, aes(time)) +
  geom_point(aes(y = observed, colour = "observed"), shape = 16) +
  geom_line(aes(y = true, colour = "true"), linewidth = 1) +
  geom_line(aes(y = fitted, colour = "fitted"), linewidth = 1) +
  scale_colour_manual(
    values = c(observed = "black", true = "grey60", fitted = "#D55E00")
  ) +
  labs(x = "Time (days)", y = "Weekly new infections", colour = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

## One parameter vs many

- `optimize()` is 1D only. For several parameters use
  `optim(pars, fn, method = "L-BFGS-B", lower = ..., upper = ...)`
- Bounds can be genuinely broad (here beta in 1e-6 to 1e-3). You do not
  need to know the neighbourhood of the truth — broad bounds plus
  several starts
- Parameters on very different scales (beta ~ 1e-4, a time ~ 40) confuse
  the default step sizes — set `control = list(parscale = ...)`
- Fit to incidence, not prevalence: incidence is what surveillance
  reports, and `cum_inc` + `diff()` gives it exactly
- Least squares on counts is a stand-in for a Poisson likelihood — fine
  for exploring, not for intervals

## Part 3: a regime change

Beta halves at a known changeover — `"constant"` interpolation holds each
knot's value until the next knot, which is exactly a step:

```{r}
sir_step <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I
  beta <- interpolate(beta_times, beta_vals, "constant")
  beta_times[] <- user()   # regime change times
  beta_vals[] <- user()    # beta in each regime
  dim(beta_times) <- user()
  dim(beta_vals) <- user()
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})
```

## Fake data with a bend at day 40

```{r}
#| fig-height: 4.2
out_step <- sir_step$new(beta_times = c(0, 40, 150),
                         beta_vals = c(2e-4, 1e-4, 1e-4))$run(week_times)
inc_step <- diff(out_step[, "cum_inc"])
obs_step <- rpois(length(inc_step), inc_step)
obs_step_df <- tibble(time = week_times[-1], obs_step = obs_step)
ggplot(obs_step_df, aes(time, obs_step)) +
  geom_point(shape = 16) +
  geom_vline(xintercept = 40, linetype = "dashed") +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

## Fitting beta1, beta2 and the change time

```{r}
sse_step <- function(pars) {   # pars = c(beta1, beta2, t_change)
  out <- sir_step$new(beta_times = c(0, pars[3], 150),
                      beta_vals = c(pars[1], pars[2], pars[2]))$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs_step)^2)
}

## broad bounds + a small grid of starts; keep the best final SSE
starts <- rbind(c(1e-4, 1e-4, 25), c(8e-4, 8e-4, 120),
                c(8e-4, 1e-5, 75), c(1e-5, 8e-4, 100))
fits <- lapply(seq_len(nrow(starts)), function(i)
  optim(starts[i, ], sse_step, method = "L-BFGS-B",
        lower = c(1e-6, 1e-6, 5), upper = c(1e-3, 1e-3, 145),
        control = list(parscale = c(1e-4, 1e-4, 10))))
sapply(fits, "[[", "value")   # final SSE per start
```

## The winning start

```{r}
fit_step <- fits[[which.min(sapply(fits, "[[", "value"))]]
fit_step$par   # truth: 2e-4, 1e-4, day 40
```

Two of the four starts land in wrong minima with about 40x worse error.
Comparing final SSE across starts — not trusting the first answer — is
what saves you. You do not need to know the truth's neighbourhood; you
need broad bounds and several starts.

## The fitted step

```{r}
#| fig-height: 3.9
out_fitstep <- sir_step$new(
  beta_times = c(0, fit_step$par[3], 150),
  beta_vals = c(fit_step$par[1], fit_step$par[2], fit_step$par[2]))$run(week_times)
fitstep_df <- tibble(
  time = week_times[-1],
  obs_step = obs_step,
  fitted = diff(out_fitstep[, "cum_inc"])
)
ggplot(fitstep_df, aes(time)) +
  geom_point(aes(y = obs_step), shape = 16) +
  geom_line(aes(y = fitted), colour = "#D55E00", linewidth = 1) +
  geom_vline(xintercept = fit_step$par[3], linetype = "dashed") +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

- Identifiability: the change time is only learnable if the curve visibly
  bends
- Local optima are real — the grid of starts is the fix

## Part 4: seasonal beta in an SIRS model

One wave ends transmission in a closed model, so let immunity wane and
drive beta seasonally:

```{r}
sirs_seasonal <- odin::odin({
  deriv(S) <- -beta * S * I + omega * R
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I - omega * R
  deriv(cum_inc) <- beta * S * I
  beta <- interpolate(beta_times, beta_vals, "linear")
  beta_times[] <- user()
  beta_vals[] <- user()
  dim(beta_times) <- user()
  dim(beta_vals) <- user()
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
  omega <- user(1 / 90)
})
```

## Two years of fake seasonal data

```{r}
#| fig-height: 3.9
beta_times <- seq(0, 730, by = 365 / 12)   # monthly knots, 2 years
beta_vals  <- 2e-4 * (1 + 0.3 * cos(2 * pi * (beta_times - 15) / 365))
times_seas <- seq(0, 730, by = 7)          # weekly output
out_seas <- sirs_seasonal$new(beta_times = beta_times,
                              beta_vals = beta_vals)$run(times_seas)
obs_seas <- rpois(length(times_seas) - 1, diff(out_seas[, "cum_inc"]))
obs_seas_df <- tibble(time = times_seas[-1], obs_seas = obs_seas)
ggplot(obs_seas_df, aes(time, obs_seas)) +
  geom_point(shape = 16, size = 1.5) +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

Peak transmission mid-January (phase = day 15), amplitude 30%.

## Fitting beta0, amplitude and phase

```{r}
sse_seas <- function(pars) {   # pars = c(beta0, amplitude, phase)
  bb <- pars[1] * (1 + pars[2] * cos(2 * pi * (beta_times - pars[3]) / 365))
  out <- sirs_seasonal$new(beta_times = beta_times,
                           beta_vals = bb)$run(times_seas)
  sum((diff(out[, "cum_inc"]) - obs_seas)^2)
}

## starts far from the truth, spread over phase
starts <- rbind(c(5e-4, 0.5, 100), c(5e-4, 0.5, 180), c(5e-4, 0.5, 300))
fits <- lapply(seq_len(nrow(starts)), function(i)
  optim(starts[i, ], sse_seas, method = "L-BFGS-B",
        lower = c(1e-6, 0, 0), upper = c(1e-3, 1, 365),
        control = list(parscale = c(1e-4, 0.1, 30))))
sapply(fits, "[[", "value")
```

## The winning start

```{r}
fit_seas <- fits[[which.min(sapply(fits, "[[", "value"))]]
fit_seas$par   # truth: 2e-4, 0.3, 15
```

Phase makes multi-start essential: one start drifts to the year boundary
(phase 365), one falls into a flat-beta minimum (amplitude 0) — only one
finds the truth. Phase is harder to fit than amplitude, because peak
timing is only visible where incidence turns; more seasons of data helps
more than denser sampling within one season.

## The fitted season

```{r}
#| fig-height: 4.2
bb_fit <- fit_seas$par[1] *
  (1 + fit_seas$par[2] * cos(2 * pi * (beta_times - fit_seas$par[3]) / 365))
out_fitseas <- sirs_seasonal$new(beta_times = beta_times,
                                 beta_vals = bb_fit)$run(times_seas)
fitseas_df <- tibble(
  time = times_seas[-1],
  obs_seas = obs_seas,
  fitted = diff(out_fitseas[, "cum_inc"])
)
ggplot(fitseas_df, aes(time)) +
  geom_point(aes(y = obs_seas), shape = 16, size = 1.5) +
  geom_line(aes(y = fitted), colour = "#D55E00", linewidth = 0.75) +
  labs(x = "Time (days)", y = "Weekly new infections") +
  theme_classic(base_size = 16)
```

## Part 5: a measured driver, made concrete

A weekly "mobility index" that dips around day 60 and recovers,
normalised to mean 1; then beta(t) = beta0 x driver(t):

```{r}
#| fig-height: 4.0
driver_times <- seq(0, 150, by = 7)
dip <- 1 - 0.6 * exp(-((driver_times - 60)^2) / (2 * 20^2))  # dip at day 60
driver <- dip / mean(dip)            # rescale to mean 1
beta_vals <- 2e-4 * driver           # beta(t) = beta0 * driver(t)
driver_df <- tibble(time = driver_times, driver = driver, beta = beta_vals)
p_driver <- ggplot(driver_df, aes(time, driver)) +
  geom_line(linewidth = 1) +
  labs(x = "Time (days)", y = "Mobility index") +
  theme_classic(base_size = 16)
p_beta_driver <- ggplot(driver_df, aes(time, beta)) +
  geom_line(colour = "#D55E00", linewidth = 1) +
  labs(x = "Time (days)", y = expression(beta(t))) +
  theme_classic(base_size = 16)
p_driver | p_beta_driver
```

## The inference variant: linear interpolation

Same SIR, but beta follows the driver smoothly between the weekly points:

```{r}
sir_driver <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I
  beta <- interpolate(beta_times, beta_vals, "linear")
  beta_times[] <- user()   # driver time points
  beta_vals[] <- user()    # beta at each time point
  dim(beta_times) <- user()
  dim(beta_vals) <- user()
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})
```

## Fitting the scaling of a known driver

```{r}
out_d <- sir_driver$new(beta_times = driver_times,
                        beta_vals = 3e-4 * driver)$run(week_times)   # truth
obs_driver <- rpois(length(week_times) - 1, diff(out_d[, "cum_inc"]))

sse_driver <- function(beta0) {   # driver known; fit only its scale
  out <- sir_driver$new(beta_times = driver_times,
                        beta_vals = beta0 * driver)$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs_driver)^2)
}
fit_driver <- optimize(sse_driver, interval = c(1e-6, 1e-3), tol = 1e-8)
fit_driver$minimum   # truth: 3e-4
```

## The fitted beta(t)

```{r}
#| fig-height: 3.9
beta_fit_df <- tibble(
  time = driver_times,
  `true beta(t)` = 3e-4 * driver,
  `fitted beta(t)` = fit_driver$minimum * driver
) |>
  pivot_longer(-time, names_to = "series", values_to = "beta") |>
  mutate(series = fct_inorder(series))
ggplot(beta_fit_df, aes(time, beta, colour = series, linetype = series)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    values = c(`true beta(t)` = "black", `fitted beta(t)` = "#D55E00")
  ) +
  scale_linetype_manual(
    values = c(`true beta(t)` = "solid", `fitted beta(t)` = "dashed")
  ) +
  coord_cartesian(ylim = c(0, 4e-4)) +
  labs(x = "Time (days)", y = expression(beta(t)),
       colour = NULL, linetype = NULL) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")
```

- This is how covariates enter transmission models: known series in, one
  coefficient out
- Several covariates = one coefficient each, fitted together
- Normalise or centre the covariate first, so the coefficient is
  interpretable — here beta0 is the average beta

## Free-form beta(t): one value per knot

```{r}
knot_times <- seq(0, 150, by = 30)   # six free betas on the Part-2 data
sse_free <- function(betas) {
  out <- sir_step$new(beta_times = knot_times,
                      beta_vals = betas)$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs)^2)
}
fit_free <- optim(rep(2e-4, 6), sse_free, method = "L-BFGS-B",
                  lower = rep(0, 6), upper = rep(5e-4, 6),
                  control = list(parscale = rep(1e-4, 6), maxit = 2000))
fit_free$par   # truth is a flat 2e-4
```

## The wiggle problem

```{r}
#| fig-height: 3.9
free_df <- tibble(time = knot_times, beta = fit_free$par)
ggplot(free_df, aes(time, beta)) +
  geom_step(linewidth = 1) +
  geom_point(shape = 16) +
  geom_hline(yintercept = 2e-4, colour = "#D55E00", linetype = "dashed") +
  coord_cartesian(xlim = c(0, 150), ylim = c(0, 4e-4)) +
  labs(x = "Time (days)", y = expression(beta)) +
  theme_classic(base_size = 16)
```

- Free knots chase noise: the truth is flat, the fit is not
- Fix: penalise roughness, e.g. add `lambda * sum(diff(betas)^2)` to the
  error
- Beyond that: odin2 + monty for proper Bayesian inference with priors
  doing the smoothing

## Choosing an approach

- **Regime change** (policy, behaviour shift): `"constant"` interpolation,
  fit the levels and the change times — broad bounds, several starts
- **Seasonal**: parametric cosine — fit beta0, amplitude, phase; spread
  the starts over phase
- **Measured driver**: pass the series as `beta_times` / `beta_vals` and
  fit a single scaling coefficient
- **Unknown shape**: free knots plus a smoothing penalty; if you need
  intervals, move to odin2 + monty

In every case: plot the series you feed in, and plot the fit against the
data before believing any number.
