---
pagetitle: "Estimating and specifying time-varying beta"
format:
  revealjs:
    theme:
      - default
      - ../theme/slides.scss
    smaller: true
    slide-number: true
    embed-resources: true
echo: true
---

```{r}
#| include: false
library(odin)
set.seed(42)
times <- seq(0, 150, by = 1)   # daily output over 150 days

## interpolate() variant of the SIR, used in Part 1 and defined in full
## in Part 3
sir_step <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I
  beta <- interpolate(beta_times, beta_vals, "constant")
  beta_times[] <- user()
  beta_vals[] <- user()
  dim(beta_times) <- user()
  dim(beta_vals) <- user()
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})
```

## Estimating and specifying time-varying beta {.title-slide}

Personal notes — not for distribution

## Three different things named "beta"

- **A model parameter** — one number per run:
  `mod <- sir_inc$new(beta = 2e-4)`
- **A scenario vector** — many runs, one per value:
  `beta_values <- c(3e-4, 2e-4, 1.5e-4)`, then `lapply`
- **A paired time series** — `beta_times` / `beta_vals` fed to
  `interpolate()`: beta varies *within* one run

The third is what these notes are about: how to specify it, and how to
estimate it from data.

## The workhorse: SIR with cumulative incidence

For fitting we need incidence, so add a `cum_inc` compartment:

```{r}
sir_inc <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I   # accumulates new infections
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  beta  <- user(2e-4)
  gamma <- user(0.1)
})
```

## The three uses in code

```{r}
## (1) one constant beta per run
out <- sir_inc$new(beta = 2e-4)$run(times)

## (2) a scenario vector: one run per value
beta_values <- c(3e-4, 2e-4, 1.5e-4)
scenario_runs <- lapply(beta_values,
                        function(beta_i) sir_inc$new(beta = beta_i)$run(times))

## (3) a paired time series: beta varies within one run
out_step <- sir_step$new(beta_times = c(0, 40, 150),
                         beta_vals = c(2e-4, 1e-4, 1e-4))$run(times)
```

## Part 2: fake data with a known beta

```{r}
week_times <- seq(0, 150, by = 7)            # weekly output
out <- sir_inc$new()$run(week_times)         # truth: beta = 2e-4
inc_true <- diff(out[, "cum_inc"])           # true weekly incidence
obs <- rpois(length(inc_true), inc_true)     # noisy "reported" data
```

```{r}
#| fig-height: 3.6
plot(week_times[-1], obs, pch = 16,
     xlab = "Time (days)", ylab = "Weekly new infections")
lines(week_times[-1], inc_true, col = "firebrick", lwd = 2)
```

## Least squares against the model

```{r}
sse <- function(beta) {
  out <- sir_inc$new(beta = beta)$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs)^2)
}

fit <- optimize(sse, interval = c(5e-5, 5e-4), tol = 1e-8)
fit$minimum   # fitted beta; truth is 2e-4
```

Spot on. Watch the tolerance: `optimize()`'s default `tol` is about 1e-4
— as big as beta itself — so without `tol = 1e-8` it "converges" to
2.2e-4, a visibly wrong curve.

## The fitted curve

```{r}
#| fig-height: 4.6
out_fit <- sir_inc$new(beta = fit$minimum)$run(week_times)
plot(week_times[-1], obs, pch = 16,
     xlab = "Time (days)", ylab = "Weekly new infections")
lines(week_times[-1], inc_true, col = "grey60", lwd = 2)
lines(week_times[-1], diff(out_fit[, "cum_inc"]), col = "firebrick", lwd = 2)
legend("topright", legend = c("observed", "true", "fitted"),
       col = c("black", "grey60", "firebrick"),
       pch = c(16, NA, NA), lty = c(NA, 1, 1), bty = "n")
```

## One parameter vs many

- `optimize()` is 1D only. For several parameters use
  `optim(pars, fn, method = "L-BFGS-B", lower = ..., upper = ...)`
- Parameters on very different scales (beta ~ 1e-4, a time ~ 40) confuse
  the default step sizes — set `control = list(parscale = ...)`
- Fit to incidence, not prevalence: incidence is what surveillance
  reports, and `cum_inc` + `diff()` gives it exactly
- Least squares on counts is a stand-in for a Poisson likelihood — fine
  for exploring, not for intervals

## Part 3: a regime change

Beta halves at a known changeover — `"constant"` interpolation holds each
knot's value until the next knot, which is exactly a step:

```{r}
sir_step <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I
  deriv(cum_inc) <- beta * S * I
  beta <- interpolate(beta_times, beta_vals, "constant")
  beta_times[] <- user()   # regime change times
  beta_vals[] <- user()    # beta in each regime
  dim(beta_times) <- user()
  dim(beta_vals) <- user()
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
})
```

## Fake data with a bend at day 40

```{r}
#| fig-height: 4.2
out_step <- sir_step$new(beta_times = c(0, 40, 150),
                         beta_vals = c(2e-4, 1e-4, 1e-4))$run(week_times)
inc_step <- diff(out_step[, "cum_inc"])
obs_step <- rpois(length(inc_step), inc_step)
plot(week_times[-1], obs_step, pch = 16,
     xlab = "Time (days)", ylab = "Weekly new infections")
abline(v = 40, lty = 2)
```

## Fitting beta1, beta2 and the change time

```{r}
sse_step <- function(pars) {   # pars = c(beta1, beta2, t_change)
  out <- sir_step$new(beta_times = c(0, pars[3], 150),
                      beta_vals = c(pars[1], pars[2], pars[2]))$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs_step)^2)
}
fit_step <- optim(c(1.5e-4, 1.5e-4, 30), sse_step, method = "L-BFGS-B",
                  lower = c(1e-5, 1e-5, 10), upper = c(5e-4, 5e-4, 120),
                  control = list(parscale = c(1e-4, 1e-4, 10)))
fit_step$par   # truth: 2e-4, 1e-4, day 40
```

## The fitted step

```{r}
#| fig-height: 3.9
out_fitstep <- sir_step$new(
  beta_times = c(0, fit_step$par[3], 150),
  beta_vals = c(fit_step$par[1], fit_step$par[2], fit_step$par[2]))$run(week_times)
plot(week_times[-1], obs_step, pch = 16,
     xlab = "Time (days)", ylab = "Weekly new infections")
lines(week_times[-1], diff(out_fitstep[, "cum_inc"]), col = "firebrick", lwd = 2)
abline(v = fit_step$par[3], lty = 2)
```

- Identifiability: the change time is only learnable if the curve visibly
  bends
- Local optima are real: a start at `(3e-4, 5e-5, 80)` lands on a bound
  with ~50x worse error — try several starts

## Part 4: seasonal beta in an SIRS model

One wave ends transmission in a closed model, so let immunity wane and
drive beta seasonally:

```{r}
sirs_seasonal <- odin::odin({
  deriv(S) <- -beta * S * I + omega * R
  deriv(I) <- beta * S * I - gamma * I
  deriv(R) <- gamma * I - omega * R
  deriv(cum_inc) <- beta * S * I
  beta <- interpolate(beta_times, beta_vals, "linear")
  beta_times[] <- user()
  beta_vals[] <- user()
  dim(beta_times) <- user()
  dim(beta_vals) <- user()
  initial(S) <- N0 - I0
  initial(I) <- I0
  initial(R) <- 0
  initial(cum_inc) <- 0
  N0    <- user(1000)
  I0    <- user(1)
  gamma <- user(0.1)
  omega <- user(1 / 90)
})
```

## Two years of fake seasonal data

```{r}
#| fig-height: 3.9
beta_times <- seq(0, 730, by = 365 / 12)   # monthly knots, 2 years
beta_vals  <- 2e-4 * (1 + 0.3 * cos(2 * pi * (beta_times - 15) / 365))
times_seas <- seq(0, 730, by = 7)          # weekly output
out_seas <- sirs_seasonal$new(beta_times = beta_times,
                              beta_vals = beta_vals)$run(times_seas)
obs_seas <- rpois(length(times_seas) - 1, diff(out_seas[, "cum_inc"]))
plot(times_seas[-1], obs_seas, pch = 16, cex = 0.6,
     xlab = "Time (days)", ylab = "Weekly new infections")
```

Peak transmission mid-January (phase = day 15), amplitude 30%.

## Fitting beta0, amplitude and phase

```{r}
sse_seas <- function(pars) {   # pars = c(beta0, amplitude, phase)
  bb <- pars[1] * (1 + pars[2] * cos(2 * pi * (beta_times - pars[3]) / 365))
  out <- sirs_seasonal$new(beta_times = beta_times,
                           beta_vals = bb)$run(times_seas)
  sum((diff(out[, "cum_inc"]) - obs_seas)^2)
}
fit_seas <- optim(c(1.5e-4, 0.2, 60), sse_seas, method = "L-BFGS-B",
                  lower = c(5e-5, 0, 0), upper = c(5e-4, 1, 365),
                  control = list(parscale = c(1e-4, 0.1, 30)))
fit_seas$par   # truth: 2e-4, 0.3, 15
```

## The fitted season

```{r}
#| fig-height: 3.9
bb_fit <- fit_seas$par[1] *
  (1 + fit_seas$par[2] * cos(2 * pi * (beta_times - fit_seas$par[3]) / 365))
out_fitseas <- sirs_seasonal$new(beta_times = beta_times,
                                 beta_vals = bb_fit)$run(times_seas)
plot(times_seas[-1], obs_seas, pch = 16, cex = 0.6,
     xlab = "Time (days)", ylab = "Weekly new infections")
lines(times_seas[-1], diff(out_fitseas[, "cum_inc"]), col = "firebrick", lwd = 1.5)
```

- Phase is harder to fit than amplitude — the peak timing is only visible
  where incidence turns
- More seasons of data helps more than denser sampling within one season

## Part 5: a measured external driver

No estimation needed — rescale the driver and pass it straight in:

```{r}
## stand-in for a measured series, e.g. a mobility index
driver <- 0.8 + 0.2 * sin(2 * pi * seq(0, 365, by = 7) / 365)

beta_times <- seq(0, 365, by = 7)
beta_vals  <- 2e-4 * driver / mean(driver)   # keep average beta at 2e-4
out_driver <- sir_step$new(beta_times = beta_times,
                           beta_vals = beta_vals)$run(seq(0, 365, by = 7))
```

Works for mobility, school terms, temperature — anything you trust as a
covariate.

## Free-form beta(t): one value per knot

```{r}
knot_times <- seq(0, 150, by = 30)   # six free betas on the Part-2 data
sse_free <- function(betas) {
  out <- sir_step$new(beta_times = knot_times,
                      beta_vals = betas)$run(week_times)
  sum((diff(out[, "cum_inc"]) - obs)^2)
}
fit_free <- optim(rep(2e-4, 6), sse_free, method = "L-BFGS-B",
                  lower = rep(0, 6), upper = rep(5e-4, 6),
                  control = list(parscale = rep(1e-4, 6), maxit = 2000))
fit_free$par   # truth is a flat 2e-4
```

## The wiggle problem

```{r}
#| fig-height: 3.9
plot(knot_times, fit_free$par, type = "s", lwd = 2, ylim = c(0, 4e-4),
     xlab = "Time (days)", ylab = expression(beta), xlim = c(0, 150))
points(knot_times, fit_free$par, pch = 16)
abline(h = 2e-4, col = "firebrick", lty = 2)
```

- Free knots chase noise: the truth is flat, the fit is not
- Fix: penalise roughness, e.g. add `lambda * sum(diff(betas)^2)` to the
  error
- Beyond that: odin2 + monty for proper Bayesian inference with priors
  doing the smoothing

## Choosing an approach

- **Regime change** (policy, behaviour shift): `"constant"` interpolation,
  fit the levels and the change times
- **Seasonal**: parametric cosine — fit beta0, amplitude, phase
- **Measured driver**: rescale and pass the series as
  `beta_times` / `beta_vals` — nothing to fit
- **Unknown shape**: free knots plus a smoothing penalty; if you need
  intervals, move to odin2 + monty

In every case: plot the series you feed in, and plot the fit against the
data before believing any number.
