---
knitr:
  opts_chunk:
    fig-width: 10
    fig-asp: 0.55
---

# Interventions: mass screening {#sec-mass-screening}

**Goal:** add test-and-isolate to the SEIR model and compare epidemic
trajectories with no screening vs mass screening starting at different times.

## The model

Start from the SEIR model and add a quarantine compartment $Q$. Mass
screening finds infected people — both exposed ($E$) and infectious ($I$) —
at rate $\delta$ and isolates them in $Q$, where they no longer transmit.
Isolated people leave quarantine at rate $\varepsilon$:

$$
\frac{dS}{dt} = -\beta S I, \qquad
\frac{dE}{dt} = \beta S I - \lambda E - \delta E,
$$

$$
\frac{dI}{dt} = \lambda E - \gamma I - \delta I, \qquad
\frac{dQ}{dt} = \delta E + \delta I - \varepsilon Q, \qquad
\frac{dR}{dt} = \gamma I + \varepsilon Q.
$$

Screening is off before `screen_time` ($\delta = 0$) and on afterwards.
Because screening shortens the time spent infectious and intercepts a
fraction $\lambda/(\lambda + \delta)$ of people before they even become
infectious, the effective reproduction number falls from
$\beta N/\gamma$ to
$\frac{\lambda}{\lambda + \delta}\,\frac{\beta N}{\gamma + \delta}$.

## odin implementation

Same pattern as school closure: one generator with a switch at
`screen_time`, cumulative incidence `inc` tracked for daily case counts.

```{r}
library(odin)
library(tidyverse)

times <- seq(0, 150, by = 1)   # daily output over 150 days

seirq <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(E) <-  beta * S * I - lambda * E - delta * E
  deriv(I) <-  lambda * E - gamma * I - delta * I
  deriv(Q) <-  delta * E + delta * I - epsilon * Q
  deriv(R) <-  gamma * I + epsilon * Q

  # cumulative incidence (differenced below to get daily cases)
  deriv(inc) <- beta * S * I

  # no screening before screen_time, screening at rate delta after
  delta <- if (t < screen_time) 0 else screen_rate

  initial(S)   <- N_ini - I_ini
  initial(E)   <- 0
  initial(I)   <- I_ini
  initial(Q)   <- 0
  initial(R)   <- 0
  initial(inc) <- 0

  N_ini <- user(5700000)
  I_ini <- user(1)

  beta        <- user(8.77e-8)  # R0 = beta * N / gamma = 2.5
  gamma       <- user(0.2)      # infectious period 5 days
  lambda      <- user(0.5)      # latent period 2 days
  screen_rate <- user(0.5)      # detected & isolated in ~2 days once on
  epsilon     <- user(0.2)      # 5 days in quarantine
  screen_time <- user(1e6)      # default: never screen
})

run_scenario <- function(screen_time, label) {
  seirq$new(screen_time = screen_time)$run(times) %>%
    as.data.frame() %>%
    mutate(scenario = label)
}
```

## Part 1: screening from day 0

Compare the uncontrolled epidemic against one where mass screening runs from
the very start:

```{r}
df_day0 <- bind_rows(
  run_scenario(1e6, "Never"),
  run_scenario(0,   "From day 0")
) %>%
  mutate(scenario = factor(scenario,
                           levels = c("Never", "From day 0")))
```

```{r}
#| fig-cap: "Daily incidence, no screening vs screening from day 0"
df_day0 %>%
  group_by(scenario) %>%
  mutate(new_cases = c(NA, diff(inc))) %>%
  ungroup() %>%
  filter(t > 0) %>%
  ggplot(aes(x = t, y = new_cases, linetype = scenario)) +
  geom_line(linewidth = 1, colour = "#D55E00") +
  xlab("Time (days)") +
  ylab("Daily new infections\n") +
  labs(linetype = "Screening") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

```{r}
#| fig-cap: "Total S, E, I, Q, R, no screening vs screening from day 0"
df_day0 %>%
  pivot_longer(c(S, E, I, Q, R), names_to = "compartment",
               values_to = "people") %>%
  ggplot(aes(x = t, y = people, colour = compartment, linetype = scenario)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("S" = "#0072B2", "E" = "#E69F00",
                                 "I" = "#D55E00", "Q" = "#CC79A7",
                                 "R" = "#009E73"),
                      limits = c("S", "E", "I", "Q", "R")) +
  xlab("Time (days)") +
  ylab("Total number of individuals\n") +
  labs(colour = NULL, linetype = "Screening") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

With screening from day 0 the effective reproduction number falls from 2.5 to
$\frac{0.5}{0.5 + 0.5}\times\frac{0.5}{0.2 + 0.5} \approx 0.36$ — the
epidemic never takes off.

## Part 2: screening from day 30

In practice screening ramps up after the epidemic is underway. Here it starts
on day 30, compared with never screening:

```{r}
screen_day <- 30

df_mid <- bind_rows(
  run_scenario(1e6, "Never"),
  run_scenario(screen_day, paste0("From day ", screen_day))
) %>%
  mutate(scenario = factor(scenario,
                           levels = c("Never",
                                      paste0("From day ", screen_day))))
```

```{r}
#| fig-cap: "Daily incidence, no screening vs screening from day 30"
df_mid %>%
  group_by(scenario) %>%
  mutate(new_cases = c(NA, diff(inc))) %>%
  ungroup() %>%
  filter(t > 0) %>%
  ggplot(aes(x = t, y = new_cases, linetype = scenario)) +
  geom_line(linewidth = 1, colour = "#D55E00") +
  geom_vline(xintercept = screen_day, linetype = "dotted", colour = "grey50",
             linewidth = 1) +
  xlab("Time (days)") +
  ylab("Daily new infections\n") +
  labs(linetype = "Screening") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

```{r}
#| fig-cap: "Total S, E, I, Q, R, no screening vs screening from day 30"
df_mid %>%
  pivot_longer(c(S, E, I, Q, R), names_to = "compartment",
               values_to = "people") %>%
  ggplot(aes(x = t, y = people, colour = compartment, linetype = scenario)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = screen_day, linetype = "dotted", colour = "grey50",
             linewidth = 1) +
  scale_colour_manual(values = c("S" = "#0072B2", "E" = "#E69F00",
                                 "I" = "#D55E00", "Q" = "#CC79A7",
                                 "R" = "#009E73"),
                      limits = c("S", "E", "I", "Q", "R")) +
  xlab("Time (days)") +
  ylab("Total number of individuals\n") +
  labs(colour = NULL, linetype = "Screening") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

Once screening switches on, $E$ and $I$ fall almost immediately — detected
cases accumulate in $Q$, where they recover without transmitting. Note the
contrast with school closure: screening removes people from the infectious
pool directly, so it bends the curve down even after the peak, while a
mixing reduction only slows new infections.
