---
knitr:
  opts_chunk:
    fig-width: 10
    fig-asp: 0.55
---

# Interventions: mass screening {#sec-mass-screening}

**Goal:** add test-and-isolate to the SEIR model and compare epidemic
trajectories with no screening vs mass screening starting at different times.

## The model

Start from the SEIR model and add a quarantine compartment $Q$. Mass
screening finds infected people — both exposed ($E$) and infectious ($I$) —
at rate $\delta$ and isolates them in $Q$, where they no longer transmit.
Isolated people leave quarantine at rate $\varepsilon$:

$$
\frac{dS}{dt} = -\beta S I, \qquad
\frac{dE}{dt} = \beta S I - \lambda E - \delta E,
$$

$$
\frac{dI}{dt} = \lambda E - \gamma I - \delta I, \qquad
\frac{dQ}{dt} = \delta E + \delta I - \varepsilon Q, \qquad
\frac{dR}{dt} = \gamma I + \varepsilon Q.
$$

Screening is off before `screen_time` ($\delta = 0$) and on afterwards.
Because screening shortens the time spent infectious and intercepts a
fraction $\lambda/(\lambda + \delta)$ of people before they even become
infectious, the effective reproduction number falls from
$\beta N/\gamma$ to
$\frac{\lambda}{\lambda + \delta}\,\frac{\beta N}{\gamma + \delta}$.
With the values below ($\delta = 0.1$, i.e. detection and isolation about
10 days after infection) this gives
$\frac{0.5}{0.6}\times\frac{0.5}{0.3} \approx 1.4$ — slower growth, but
still above 1. Detection faster than about every 6 days
($\delta \gtrsim 0.17$) would push it below 1 and stop the epidemic
outright.

## odin implementation

Same pattern as school closure: one generator with a switch at
`screen_time`, cumulative incidence `inc` tracked for daily case counts.
We seed 1000 infectious people and run 300 days, so every scenario shows
its full course — rise, peak and decline to stability.

```{r}
library(odin)
library(tidyverse)

times <- seq(0, 300, by = 1)   # daily output over 300 days

seirq <- odin::odin({
  deriv(S) <- -beta * S * I
  deriv(E) <-  beta * S * I - lambda * E - delta * E
  deriv(I) <-  lambda * E - gamma * I - delta * I
  deriv(Q) <-  delta * E + delta * I - epsilon * Q
  deriv(R) <-  gamma * I + epsilon * Q

  # cumulative incidence (differenced below to get daily cases)
  deriv(inc) <- beta * S * I

  # no screening before screen_time, screening at rate delta after
  delta <- if (t < screen_time) 0 else screen_rate

  initial(S)   <- N_ini - I_ini
  initial(E)   <- 0
  initial(I)   <- I_ini
  initial(Q)   <- 0
  initial(R)   <- 0
  initial(inc) <- 0

  N_ini <- user(5700000)
  I_ini <- user(1000)   # start with 1000 infectious, so dynamics are visible

  beta        <- user(8.77e-8)  # R0 = beta * N / gamma = 2.5
  gamma       <- user(0.2)      # infectious period 5 days
  lambda      <- user(0.5)      # latent period 2 days
  screen_rate <- user(0.1)      # detected & isolated in ~10 days once on
  epsilon     <- user(0.2)      # 5 days in quarantine
  screen_time <- user(1e6)      # default: never screen
})

run_scenario <- function(screen_time, label) {
  seirq$new(screen_time = screen_time)$run(times) %>%
    as.data.frame() %>%
    mutate(scenario = label)
}
```

## Part 1: screening from day 0

Compare the uncontrolled epidemic against one where mass screening runs from
the very start:

```{r}
df_day0 <- bind_rows(
  run_scenario(1e6, "Never"),
  run_scenario(0,   "From day 0")
) %>%
  mutate(scenario = factor(scenario,
                           levels = c("Never", "From day 0")))
```

```{r}
#| fig-cap: "Daily incidence, no screening vs screening from day 0"
df_day0 %>%
  group_by(scenario) %>%
  mutate(new_cases = c(NA, diff(inc))) %>%
  ungroup() %>%
  filter(t > 0) %>%
  ggplot(aes(x = t, y = new_cases, linetype = scenario)) +
  geom_line(linewidth = 1, colour = "#D55E00") +
  xlab("Time (days)") +
  ylab("Daily new infections\n") +
  labs(linetype = "Screening") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

```{r}
#| fig-cap: "Total S, E, I, Q, R, no screening vs screening from day 0"
df_day0 %>%
  pivot_longer(c(S, E, I, Q, R), names_to = "compartment",
               values_to = "people") %>%
  ggplot(aes(x = t, y = people, colour = compartment, linetype = scenario)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("S" = "#0072B2", "E" = "#E69F00",
                                 "I" = "#D55E00", "Q" = "#CC79A7",
                                 "R" = "#009E73"),
                      limits = c("S", "E", "I", "Q", "R")) +
  xlab("Time (days)") +
  ylab("Total number of individuals\n") +
  labs(colour = NULL, linetype = "Screening") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

With screening from day 0 the effective reproduction number falls from 2.5 to
about 1.4 — the epidemic still grows, but far more slowly. The peak arrives
around day 94 at roughly 140,000 infectious, versus 940,000 on day 52 with
no screening, and the final size falls from 5.1 to 2.9 million. Note the
isolated compartment $Q$ swelling as cases are detected, then emptying as
they leave quarantine.

## Part 2: screening from day 30

In practice screening ramps up after the epidemic is underway. Here it starts
on day 30, compared with never screening:

```{r}
screen_day <- 30

df_mid <- bind_rows(
  run_scenario(1e6, "Never"),
  run_scenario(screen_day, paste0("From day ", screen_day))
) %>%
  mutate(scenario = factor(scenario,
                           levels = c("Never",
                                      paste0("From day ", screen_day))))
```

```{r}
#| fig-cap: "Daily incidence, no screening vs screening from day 30"
df_mid %>%
  group_by(scenario) %>%
  mutate(new_cases = c(NA, diff(inc))) %>%
  ungroup() %>%
  filter(t > 0) %>%
  ggplot(aes(x = t, y = new_cases, linetype = scenario)) +
  geom_line(linewidth = 1, colour = "#D55E00") +
  geom_vline(xintercept = screen_day, linetype = "dotted", colour = "grey50",
             linewidth = 1) +
  xlab("Time (days)") +
  ylab("Daily new infections\n") +
  labs(linetype = "Screening") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

```{r}
#| fig-cap: "Total S, E, I, Q, R, no screening vs screening from day 30"
df_mid %>%
  pivot_longer(c(S, E, I, Q, R), names_to = "compartment",
               values_to = "people") %>%
  ggplot(aes(x = t, y = people, colour = compartment, linetype = scenario)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = screen_day, linetype = "dotted", colour = "grey50",
             linewidth = 1) +
  scale_colour_manual(values = c("S" = "#0072B2", "E" = "#E69F00",
                                 "I" = "#D55E00", "Q" = "#CC79A7",
                                 "R" = "#009E73"),
                      limits = c("S", "E", "I", "Q", "R")) +
  xlab("Time (days)") +
  ylab("Total number of individuals\n") +
  labs(colour = NULL, linetype = "Screening") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

Once screening switches on, growth slows visibly — the peak is cut from
940,000 to about 205,000 and the final size from 5.1 to 3.1 million — but
with $\delta = 0.1$ the effective $R$ stays above 1, so the epidemic still
runs its course before settling. Faster detection (shorter $1/\delta$,
beyond about $\delta = 0.17$) is what would bend the curve down and stop it
outright. Compare school closure: both interventions here leave $R > 1$ and
so buy time and lower the peak; stopping an epidemic requires pushing the
effective $R$ below 1.
