---
knitr:
  opts_chunk:
    fig-width: 10
    fig-asp: 0.55
---

# Interventions: school closure {#sec-school-closure}

**Goal:** extend the SIR model to two age groups — children and adults — and
use it to compare epidemic trajectories with schools open vs closed.

## The model

The population is split into children ($c$) with $N_c = 700\,000$ and adults
($a$) with $N_a = 5\,000\,000$. $\beta_{ij}$ is the per-capita rate at which
one infectious person in group $j$ infects a susceptible in group $i$, so the
force of infection on each group mixes infectious people from both groups:

$$
\frac{dS_c}{dt} = -(\beta_{cc} I_c + \beta_{ac} I_a)\, S_c, \qquad
\frac{dI_c}{dt} = (\beta_{cc} I_c + \beta_{ac} I_a)\, S_c - \gamma I_c,
$$

$$
\frac{dS_a}{dt} = -(\beta_{ca} I_c + \beta_{aa} I_a)\, S_a, \qquad
\frac{dI_a}{dt} = (\beta_{ca} I_c + \beta_{aa} I_a)\, S_a - \gamma I_a,
$$

with $\frac{dR_c}{dt} = \gamma I_c$, $\frac{dR_a}{dt} = \gamma I_a$ and
$\gamma = 1/5$ per day (mean infectious period of 5 days) in both groups.

At the start of the epidemic one infectious child infects
$(\beta_{cc} S_c + \beta_{ca} S_a)/\gamma$ people and one infectious adult
infects $(\beta_{ac} S_c + \beta_{aa} S_a)/\gamma$. Closing schools mostly
cuts child-to-child mixing ($\beta_{cc}$); the other rates barely move:

| mixing rate    | schools open | schools closed |
|----------------|-------------:|---------------:|
| $\beta_{cc}$   | 0.000 000 800 | 0.000 000 300 |
| $\beta_{ca}$   | 0.000 000 060 | 0.000 000 066 |
| $\beta_{ac}$   | 0.000 000 200 | 0.000 000 220 |
| $\beta_{aa}$   | 0.000 000 050 | 0.000 000 045 |
| 1 child infects| 4.3           | 2.7           |
| 1 adult infects| 1.95          | 1.90          |

## odin implementation

One generator covers every scenario: each mixing rate switches from its
"schools open" value to its "schools closed" value at `close_time`. Setting
`close_time = 0` closes schools from day 0; setting it beyond the simulation
horizon keeps them open. Two extra compartments `inc_c` and `inc_a` accumulate
new infections so daily incidence can be recovered by differencing.

```{r}
library(odin)
library(tidyverse)

times <- seq(0, 150, by = 1)   # daily output over 150 days

sir_age <- odin::odin({
  deriv(S_c) <- -(beta_cc * I_c + beta_ac * I_a) * S_c
  deriv(I_c) <-  (beta_cc * I_c + beta_ac * I_a) * S_c - gamma * I_c
  deriv(R_c) <-   gamma * I_c

  deriv(S_a) <- -(beta_ca * I_c + beta_aa * I_a) * S_a
  deriv(I_a) <-  (beta_ca * I_c + beta_aa * I_a) * S_a - gamma * I_a
  deriv(R_a) <-   gamma * I_a

  # cumulative incidence (differenced below to get daily cases)
  deriv(inc_c) <- (beta_cc * I_c + beta_ac * I_a) * S_c
  deriv(inc_a) <- (beta_ca * I_c + beta_aa * I_a) * S_a

  # schools open until close_time, closed afterwards
  beta_cc <- if (t < close_time) beta_cc_open else beta_cc_closed
  beta_ca <- if (t < close_time) beta_ca_open else beta_ca_closed
  beta_ac <- if (t < close_time) beta_ac_open else beta_ac_closed
  beta_aa <- if (t < close_time) beta_aa_open else beta_aa_closed

  initial(S_c) <- N_c - I_ini
  initial(I_c) <- I_ini
  initial(R_c) <- 0
  initial(S_a) <- N_a - I_ini
  initial(I_a) <- I_ini
  initial(R_a) <- 0
  initial(inc_c) <- 0
  initial(inc_a) <- 0

  N_c   <- user(700000)
  N_a   <- user(5000000)
  I_ini <- user(1)      # one infectious seed in each age group
  gamma <- user(0.2)    # infectious period 5 days

  # mixing rates, schools open / closed
  beta_cc_open <- user(8e-7)
  beta_ca_open <- user(6e-8)
  beta_ac_open <- user(2e-7)
  beta_aa_open <- user(5e-8)

  beta_cc_closed <- user(3e-7)
  beta_ca_closed <- user(6.6e-8)
  beta_ac_closed <- user(2.2e-7)
  beta_aa_closed <- user(4.5e-8)

  close_time <- user(1e6)  # default: schools never close
})
```

A small helper runs one scenario and labels the output:

```{r}
run_scenario <- function(close_time, label) {
  sir_age$new(close_time = close_time)$run(times) %>%
    as.data.frame() %>%
    mutate(scenario = label)
}
```

## Part 1: schools closed from day 0

Compare the epidemic with schools open throughout against one where schools
are closed from the very start:

```{r}
df_day0 <- bind_rows(
  run_scenario(1e6, "Open"),
  run_scenario(0,   "Closed (day 0)")
) %>%
  mutate(scenario = factor(scenario,
                           levels = c("Open", "Closed (day 0)")))
```

Daily new infections in each age group — the view where school closure
actually bites:

```{r}
#| fig-cap: "Daily incidence by age group, schools open vs closed from day 0"
df_day0 %>%
  group_by(scenario) %>%
  mutate(new_c = c(NA, diff(inc_c)), new_a = c(NA, diff(inc_a))) %>%
  ungroup() %>%
  filter(t > 0) %>%
  pivot_longer(c(new_c, new_a), names_to = "group", values_to = "cases") %>%
  mutate(group = recode(group, new_c = "Children", new_a = "Adults")) %>%
  ggplot(aes(x = t, y = cases, colour = group, linetype = scenario)) +
  geom_line(linewidth = 1) +
  xlab("Time (days)") +
  ylab("Daily new infections\n") +
  scale_colour_manual(values = c("Children" = "#E69F00", "Adults" = "#56B4E9")) +
  labs(colour = "Age group", linetype = "Scenario") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

And the overall S/I/R trajectories:

```{r}
#| fig-cap: "Total S, I, R, schools open vs closed from day 0"
df_day0 %>%
  mutate(S = S_c + S_a, I = I_c + I_a, R = R_c + R_a) %>%
  pivot_longer(c(S, I, R), names_to = "compartment", values_to = "people") %>%
  ggplot(aes(x = t, y = people, colour = compartment, linetype = scenario)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("S" = "#0072B2", "I" = "#D55E00",
                                 "R" = "#009E73"),
                      limits = c("S", "I", "R")) +
  xlab("Time (days)") +
  ylab("Total number of individuals\n") +
  labs(colour = "Compartment", linetype = "Scenario") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

Closing schools lowers and delays the peak. The effect is strongest in
children — exactly where $\beta_{cc}$ is cut from $8\times10^{-7}$ to
$3\times10^{-7}$ — but adults benefit too, because fewer infected children
means fewer infections spilling over through $\beta_{ca}$.

## Part 2: closing schools mid-epidemic

In practice interventions start partway through. Here schools close on day 15,
compared with never closing:

```{r}
close_day <- 15

df_mid <- bind_rows(
  run_scenario(1e6, "Open"),
  run_scenario(close_day, paste0("Closed (day ", close_day, ")"))
) %>%
  mutate(scenario = factor(scenario,
                           levels = c("Open",
                                      paste0("Closed (day ", close_day, ")"))))
```

```{r}
#| fig-cap: "Daily incidence by age group, schools open vs closed on day 15"
df_mid %>%
  group_by(scenario) %>%
  mutate(new_c = c(NA, diff(inc_c)), new_a = c(NA, diff(inc_a))) %>%
  ungroup() %>%
  filter(t > 0) %>%
  pivot_longer(c(new_c, new_a), names_to = "group", values_to = "cases") %>%
  mutate(group = recode(group, new_c = "Children", new_a = "Adults")) %>%
  ggplot(aes(x = t, y = cases, colour = group, linetype = scenario)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = close_day, linetype = "dashed", colour = "grey",
             linewidth = 1) +
  xlab("Time (days)") +
  ylab("Daily new infections\n") +
  scale_colour_manual(values = c("Children" = "#E69F00", "Adults" = "#56B4E9")) +
  labs(colour = "Age group", linetype = "Scenario") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

```{r}
#| fig-cap: "Total S, I, R, schools open vs closed on day 15"
df_mid %>%
  mutate(S = S_c + S_a, I = I_c + I_a, R = R_c + R_a) %>%
  pivot_longer(c(S, I, R), names_to = "compartment", values_to = "people") %>%
  ggplot(aes(x = t, y = people, colour = compartment, linetype = scenario)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = close_day, linetype = "dashed", colour = "grey",
             linewidth = 1) +
  scale_colour_manual(values = c("S" = "#0072B2", "I" = "#D55E00",
                                 "R" = "#009E73"),
                      limits = c("S", "I", "R")) +
  xlab("Time (days)") +
  ylab("Total number of individuals\n") +
  labs(colour = "Compartment", linetype = "Scenario") +
  theme_classic(base_size = 20) +
  theme(legend.position = "bottom")
```

Incidence drops immediately after day 15, and because closure happens early in
the growth phase, most infections are averted — the trajectory is much closer
to closing from day 0 than to never closing. Timing matters as much as the
size of the mixing change.
